---
title: "LAMPRA Paper Figures — Fig2 & FigS2 & FigS3"
author: "cbiar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    fig_width: 10
    fig_height: 10
  pdf_document:
    toc: true
    toc_depth: '3'
editor_options:
  chunk_output_type: console
---

<!--
OVERVIEW
========
  Fig 2a - LAMPRA activity as a function of upstream enhancer #
  Fig 2b - LAMPRA activity per element copy number
  Fig 2c - Relative positional activity of e-NMU(iii), (iv), and Insu(i)
  Fig 2d - Bar plot of LAMPRA activity for upstream vs. downstream Insu(i)
  Fig 2e - Bar plot of LAMPRA activity for upstream vs. downstream Insu(i) for all sCRL architectures
  Fig 2f - Bar chart of Pearson r scores for all models
  Fig 2g - Scatterplot of random forest predicted vs. actual activity
  Fig 2h - Downsampling analysis curve
  
  Fig S2a - Violin plot of LAMPRA activity as a function of upstream enhancer #
  Fig S2b - LAMPRA activity per element copy number for active and inactive elements
  Fig S2c - Orientation sensitivity (Forward − Reverse) by slot
  Fig S2d - Scatterplot of mean LAMPRA score by slot and orientation

  Fig S3a - Violin plot of LAMPRA activity by insulator copy number
  Fig S3b - Violin plot of LAMPRA activity for upstream vs. downstream Insu(i)
  Fig S3c - Violin plot of LAMPRA activity for upstream vs. downstream Insu(i) for all sCRL architectures
  Fig S3d - Scatterplot of mean LAMPRA score by slot and orientation for Insu(i) only

DEPENDENCIES (CRAN)
  tidyverse, scales, RColorBrewer, cowplot, sysfonts, showtext, here.
  Font: macOS Arial paths used — edit font_add() for Linux/Windows.
-->

```{r setup, include=FALSE}
FIG_DIR <- file.path(here::here(), "figures-fig2-figS2-figS3")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 10,
  dpi  = 150,
  dev  = c("pdf", "png"),
  fig.path = paste0(FIG_DIR, "/")
)
```

# Setup

## Libraries

```{r libraries}
library(tidyverse)  
library(scales)       
library(RColorBrewer) 
library(cowplot)      
library(sysfonts)     
library(showtext)     
library(egg)          
library(gtable)
library(grid)
library(ggbeeswarm)
```

## Font & Theme

```{r font-theme}
# macOS Arial paths. On Linux/Windows point font_add() at a local .ttf,
# or remove this block entirely — plots fall back to the device default font.
font_add("Arial", regular    = "/Library/Fonts/Arial.ttf",
                  bold       = "/Library/Fonts/Arial Bold.ttf",
                  italic     = "/Library/Fonts/Arial Italic.ttf",
                  bolditalic = "/Library/Fonts/Arial Bold Italic.ttf")
showtext_auto()

FONT <- "Arial"

# ── Publication theme ─────────────────────────────────────────────────────────
pub_theme <- function() {
  theme_cowplot(font_size = 8) %+replace%
    theme(
      strip.background   = element_rect(color = "black", fill = "grey90",
                                        linetype = "solid", linewidth = 0.25),
      text               = element_text(size = 8, family = FONT),
      axis.text.x        = element_text(size = 8),
      axis.text.y        = element_text(size = 8, hjust = 1,
                                        margin = margin(r = 2)),
      axis.title.x      = element_text(size = 9, margin = margin(t = -1)),
      axis.title.y      = element_text(size = 9, angle = 90, margin = margin(r = -1)),
      axis.line          = element_line(linewidth = 0.25),
      strip.text         = element_text(family = FONT, size = 8, face = "bold",
                                        margin = margin(b = 2.5, t = 2.5)),
      legend.text        = element_text(size = 8),
      legend.title       = element_text(size = 8),
      legend.position    = "right",
      legend.title.align = 0.5,
      plot.tag           = element_text(size = 14, face = "bold"),
      plot.title         = element_text(hjust = 0.5, face = "plain", size = 12,
                                        margin = margin(b = 10))
    )
}
theme_set(pub_theme())

# ── Panel dimension constants ─────────────────────────────────────────────────
# NOTE: these are per-PANEL sizes (the plotting area), not total figure size —
# save_fig() uses egg::set_panel_size() so panel width/height stays identical
# across figures regardless of legend/label footprint.
PANEL_W      <- unit(2.3, "in")
PANEL_H      <- unit(2.3, "in")
PANEL_H      <- PANEL_H * 0.9

ANNOT_SIZE <- 8 / .pt   # matches theme's 10pt in ggplot2 size units

# Single accent color for the learning curves (Set2 teal — "e-NMU(iii)F").
COLOR_LINE <- "black"

# ── save_fig() ────────────────────────────────────────────────────────────────
# Writes to FIG_DIR via quartz's native PDF device (real embedded Arial text,
# not showtext-outlined paths — this is what keeps text selectable/editable
# as whole words in Illustrator).
#
# Backward-compatible: existing calls using width=/height= (old ggsave-style,
# total figure inches) still work — they're mapped to panel_width/panel_height.
save_fig <- function(name, plot = last_plot(),
                     width = NULL, height = NULL,
                     panel_width  = width  %||% PANEL_W,
                     panel_height = height %||% PANEL_H,
                     dpi = 300, ext = "pdf", ...) {
  fpath <- file.path(FIG_DIR, paste0(name, ".", ext))

  if (!inherits(panel_width,  "unit")) panel_width  <- unit(panel_width,  "in")
  if (!inherits(panel_height, "unit")) panel_height <- unit(panel_height, "in")

  sized   <- egg::set_panel_size(plot, width = panel_width, height = panel_height)
  total_w <- grid::convertWidth( gtable::gtable_width(sized),  "in", valueOnly = TRUE)
  total_h <- grid::convertHeight(gtable::gtable_height(sized), "in", valueOnly = TRUE)

  if (ext == "pdf") {
    quartz(type = "pdf", file = fpath,
           width = total_w, height = total_h, family = "Arial")
    grid::grid.draw(sized)
    dev.off()
  } else if (ext == "png") {
    quartz(type = "png", file = fpath,
           width = total_w * dpi, height = total_h * dpi, family = "Arial")
    grid::grid.draw(sized)
    dev.off()
  }

  if (file.exists(fpath) && file.size(fpath) > 0) {
    message(sprintf("Saved: %s  [%.2f × %.2f in]", fpath, total_w, total_h))
  } else {
    warning(sprintf("FAILED to write: %s", fpath))
  }
  invisible(fpath)
}

# ── r2_score() ────────────────────────────────────────────────────────────────
# coefficient of determination - sklearn-compatible R² = 1 − SS_res / SS_tot. Can be negative for subsets.
r2_score <- function(actual, predicted) {
  ss_res <- sum((actual - predicted)^2)
  ss_tot <- sum((actual - mean(actual))^2)
  if (ss_tot == 0) return(NA_real_)
  1 - ss_res / ss_tot
}
```

## Input Paths

| Variable       | File                                                        
|----------------|-------------------------------------------------------------
| `RAW_DATA`     | `data/raw/v2_longMPRA_scores_with_orientation_20260127.txt` 
| `RFXGB_FULL`   | `data/categorical/full_processed_categorical.csv`           
| `HOLDOUT_EVAL` | `data/categorical/holdout_eval_combined.csv`               
| `DOWNSAMPLE`   | `data/categorical/downsampling_metrics.csv`                 

```{r input-paths}
PROJECT <- here::here()

RAW_DATA <- file.path(PROJECT,
  "data/raw/v2_longMPRA_scores_with_orientation_20260127.txt")

RFXGB_FULL <- file.path(PROJECT,
  "data/categorical/full_processed_categorical.csv")

HOLDOUT_EVAL <- file.path(PROJECT,
  "data/categorical/holdout_eval_combined.csv")

DOWNSAMPLE <- file.path(PROJECT,
  "data/categorical/downsampling_metrics.csv")
```

## Categorical Code → Label Maps

`full_processed_categorical.csv` stores element identities as integers
(sklearn label-encoded, alphabetical order). `decode_categorical_codes()`
converts those integers back to string labels in place.

```{r categorical-decode}
# sklearn alphabetical order of the 11 insert identities → integer code
ID_CODE_MAP <- c(
  "0"  = "dinucleotide_shuffled_i",
  "1"  = "dinucleotide_shuffled_ii",
  "2"  = "dinucleotide_shuffled_iii",
  "3"  = "dinucleotide_shuffled_iv",
  "4"  = "dinucleotide_shuffled_v",
  "5"  = "eNMU_region_i",
  "6"  = "eNMU_region_ii",
  "7"  = "eNMU_region_iii",
  "8"  = "eNMU_region_iv",
  "9"  = "eNMU_region_v",
  "10" = "synthetic_insulator"
)
ORI_CODE_MAP <- c("0" = "forward", "1" = "reverse")

decode_categorical_codes <- function(df, slots = 1:5) {
  for (s in slots) {
    id_code  <- paste0("insert_", s, "_id_code")
    ori_code <- paste0("insert_", s, "_ori_code")
    if (id_code  %in% names(df))
      df[[paste0("insert_", s, "_id")]]  <-
        unname(ID_CODE_MAP[as.character(df[[id_code]])])
    if (ori_code %in% names(df))
      df[[paste0("insert_", s, "_ori")]] <-
        unname(ORI_CODE_MAP[as.character(df[[ori_code]])])
  }
  df
}
```

## Color Palette

```{r colors}
# RColorBrewer Set2 (8 colors) with semantic names
SET2 <- brewer.pal(8, "Set2")
names(SET2) <- c("teal", "orange", "lavender", "pink",
                 "yellow_green", "yellow", "tan", "gray")

COLOR_INSULATOR    <- SET2["orange"]
COLOR_DINUC_SHUF   <- SET2["lavender"]
COLOR_ENMU_GENERIC <- SET2["teal"]
COLOR_FALLBACK     <- "#000000"
COLOR_IIIF         <- SET2["teal"]
COLOR_OTHER        <- "gray"

# e-NMU ramp: light teal → Set2 teal (e-NMU(iii) anchor) → dark teal
ENMU_LIGHT <- "#DCEFE7"; ENMU_BASE <- "#66C2A5"; ENMU_DARK <- "#0A4A33"
enmu_palette <- colorRampPalette(c(ENMU_LIGHT, ENMU_BASE, ENMU_DARK))(5)
ENMU_OVERRIDES <- c(NA, NA, NA, NA, NA)
enmu_palette   <- ifelse(is.na(ENMU_OVERRIDES), enmu_palette, ENMU_OVERRIDES)

# Ctrl ramp: light lavender → Set2 lavender (Ctrl(iii) anchor) → navy
CTRL_LIGHT <- "#E5EAF4"; CTRL_BASE <- "#8DA0CB"; CTRL_DARK <- "#1F2C5C"
ctrl_palette <- colorRampPalette(c(CTRL_LIGHT, CTRL_BASE, CTRL_DARK))(5)
CTRL_OVERRIDES <- c(NA, NA, NA, NA, NA)
ctrl_palette   <- ifelse(is.na(CTRL_OVERRIDES), ctrl_palette, CTRL_OVERRIDES)

# Named color map for all 11 element types
elem_colors <- c(
  setNames(enmu_palette, c("e-NMU(i)","e-NMU(ii)","e-NMU(iii)","e-NMU(iv)","e-NMU(v)")),
  setNames(ctrl_palette, c("Ctrl(i)", "Ctrl(ii)", "Ctrl(iii)", "Ctrl(iv)", "Ctrl(v)")),
  c("Insu(i)" = unname(COLOR_INSULATOR))
)

# Relative-activity plots exclude Ctrl from the line colors
# (Ctrl is the baseline denominator, not a plotted line).
elem_colors_rel <- elem_colors[!grepl("^Ctrl\\(", names(elem_colors))]

get_part_color <- function(name) {
  case_when(
    str_detect(name, "dinucleotide_shuffled")   ~ unname(COLOR_DINUC_SHUF),
    str_detect(name, "synthetic_insulator")     ~ unname(COLOR_INSULATOR),
    TRUE                                        ~ unname(COLOR_FALLBACK)
  )
}
```

---

# Data Loading

## Raw LAMPRA scores
```{r load-raw-data}
df_raw <- read.table(RAW_DATA, header = TRUE, sep = "\t",
                     stringsAsFactors = FALSE)

df <- df_raw %>%
  mutate(
    # S5 label used to stratify Fig 2a and highlight the iiiF subset.
    S5 = if_else(insert_5 == "eNMU_region_iii_forward", "e-NMU(iii)F", "Other"),
    # Pseudocount 0.05 shifts the distribution away from zero before logging.
    # It is used only for visualisation / log transforms; raw avg_Rep is used
    # for all model inputs and metrics.
    avg_Rep_log05 = log10(avg_Rep + 0.05)
  )

cat(nrow(df), "constructs loaded.\n")
cat("S5 breakdown:\n"); print(table(df$S5))
```

## Build ori_long
`ori_long` pivots the five slot columns into long format: one row per
(construct × slot). Each row carries the element identity, orientation, raw
score, and log score for that slot. 
```{r build-ori-long}
ori_long <- map_dfr(1:5, function(s) {
  df %>%
    transmute(
      slot          = s,
      element_id    = .data[[paste0("insert_", s, "_id")]],
      orientation   = .data[[paste0("insert_", s, "_ori")]],
      avg_Rep       = avg_Rep,           # raw score — alt orientation plots
      avg_Rep_log05 = avg_Rep_log05      # log10(avg_Rep + 0.05)
    )
}) %>%
  filter(!is.na(avg_Rep_log05)) %>%
  mutate(
    # Convert raw element IDs to publication display labels.
    element_label = element_id %>%
      str_replace_all("eNMU_region_([ivx]+)",           "e-NMU(\\1)") %>%
      str_replace_all("dinucleotide_shuffled_([ivx]+)", "Ctrl(\\1)")   %>%
      str_replace_all("synthetic_insulator",             "Insu(i)")
  )

cat(nrow(ori_long), "element × construct × slot rows in ori_long.\n")
```

## Load categorical feature matrix
```{r load-rfxgb-full}
# full_processed_categorical.csv: one row per construct.
# Element identities are integer-coded (sklearn label encoding).
# decode_categorical_codes() appends string _id and _ori columns.
rfxgb_full <- read_csv(RFXGB_FULL, show_col_types = FALSE) %>%
  decode_categorical_codes()

cat(nrow(rfxgb_full), "rows in rfxgb_full after decoding.\n")
```

## Build iiiF slot data

### Subset definition

`df_iiiF` contains all constructs where **slot 5 = e-NMU(iii) forward** (n ≈ 1,665). 

### How `build_elements_long` works

For each of the 11 element types and each of the 4 variable slots, we pull
every iiiF construct that has that element at that slot. The resulting long
table has one row per (element type, slot, construct).

### Geometric mean
### log2 FC = log₂(geomean_A / geomean_Ctrl)

```
delta_log = mean_log_elem − mean_log_ctrl
          = log10( geomean_A ) − log10( geomean_Ctrl )
          = log10( geomean_A / geomean_Ctrl )
```

```
log2_fc = delta_log / log10(2)
        = log10( geomean_A / geomean_Ctrl ) / log10(2)
        = log2( geomean_A / geomean_Ctrl )
```

```{r build-iiiF-slot-data}
# ── iiiF-S5 subset ────────────────────────────────────────────────────────────
df_iiiF <- df %>% filter(S5 == "e-NMU(iii)F")
cat(nrow(df_iiiF), "constructs in iiiF subset.\n")

# Lookup table: all 11 element types with publication labels and family.
elements_11 <- tribble(
  ~element_id,                  ~element_label,  ~family,
  "eNMU_region_i",              "e-NMU(i)",      "e-NMU",
  "eNMU_region_ii",             "e-NMU(ii)",     "e-NMU",
  "eNMU_region_iii",            "e-NMU(iii)",    "e-NMU",
  "eNMU_region_iv",             "e-NMU(iv)",     "e-NMU",
  "eNMU_region_v",              "e-NMU(v)",      "e-NMU",
  "dinucleotide_shuffled_i",    "Ctrl(i)",        "Ctrl",
  "dinucleotide_shuffled_ii",   "Ctrl(ii)",       "Ctrl",
  "dinucleotide_shuffled_iii",  "Ctrl(iii)",      "Ctrl",
  "dinucleotide_shuffled_iv",   "Ctrl(iv)",       "Ctrl",
  "dinucleotide_shuffled_v",    "Ctrl(v)",        "Ctrl",
  "synthetic_insulator",        "Insu(i)",        "Insu"
)

# ── build_elements_long() ─────────────────────────────────────────────────────
# For each (element_type × slot S1–S4), find every construct in df_base that
# has that element at that slot. Returns a long table with one row per
# (construct, element, slot) — the denominator for all per-slot summaries.
# Note: constructs appear multiple times if they carry multiple element types,
# but each (element, slot) analysis is independent (other slots are free).
build_elements_long <- function(df_base) {
  map_dfr(seq_len(nrow(elements_11)), function(i) {
    elem <- elements_11$element_id[i]
    lbl  <- elements_11$element_label[i]
    fam  <- elements_11$family[i]
    map_dfr(1:4, function(s) {
      col <- paste0("insert_", s, "_id")
      df_base %>%
        filter(.data[[col]] == elem) %>%
        transmute(slot = s, element = lbl, family = fam, avg_Rep)
    })
  }) %>%
    mutate(
      element = factor(element, levels = elements_11$element_label),
      family  = factor(family,  levels = c("e-NMU", "Ctrl", "Insu"))
    )
}

# ── build_relative_summary() ──────────────────────────────────────────────────
# Computes, for each non-Ctrl element at each slot:
#   log2_fc = log2( geomean(element activity) / geomean(Ctrl-pool activity) )
#
# Steps (matching the algebra above):
#   1. For each slot, pool ALL Ctrl constructs (all five Ctrl variants combined)
#      → mean_log_ctrl  = log10(geomean of Ctrl-pool at that slot)
#   2. For each (element, slot), average log10 activity of constructs bearing
#      that element at that slot
#      → mean_log_elem  = log10(geomean of element A at that slot)
#   3. delta_log = mean_log_elem − mean_log_ctrl
#              = log10(geomean_A / geomean_Ctrl)
#   4. log2_fc  = delta_log / log10(2) = log2(geomean_A / geomean_Ctrl)
#   5. Error:   se_delta = sqrt(se_elem² + se_ctrl²)   [independent groups]
#              ci95     = 1.96 × se_delta              [large-sample 95% CI]
build_relative_summary <- function(elements_long) {
  elements_log <- elements_long %>%
    mutate(log_act = log10(avg_Rep + 0.05))

  # Step 1 — Ctrl-pool mean per slot (all five Ctrl variants pooled together)
  ctrl_baseline <- elements_log %>%
    filter(family == "Ctrl") %>%
    group_by(slot) %>%
    summarise(
      mean_log_ctrl = mean(log_act),        # = log10(geomean_Ctrl at this slot)
      se_log_ctrl   = sd(log_act) / sqrt(n()),
      n_ctrl        = n(),
      .groups = "drop"
    )

  # Step 2 — Element mean per (element, slot) for non-Ctrl elements
  elem_mean_log <- elements_log %>%
    filter(family != "Ctrl") %>%
    group_by(family, element, slot) %>%
    summarise(
      mean_log_elem = mean(log_act),        # = log10(geomean_elem at this slot)
      se_log_elem   = sd(log_act) / sqrt(n()),
      n_elem        = n(),
      .groups = "drop"
    )

  # Steps 3–5 — Join and compute log2 FC with 95% CI
  elem_mean_log %>%
    inner_join(ctrl_baseline, by = "slot") %>%
    mutate(
      delta_log  = mean_log_elem - mean_log_ctrl,   # log10(geomean_A/geomean_Ctrl)
      se_delta   = sqrt(se_log_elem^2 + se_log_ctrl^2),
      ci95_delta = se_delta * 1.96,
      log2_fc    = delta_log / log10(2),            # log2(geomean_A/geomean_Ctrl)
      log2_lo    = (delta_log - ci95_delta) / log10(2),
      log2_hi    = (delta_log + ci95_delta) / log10(2)
    )
}
```

## Load model outputs
```{r load-model-outputs}
# The combined file stacks two evaluation runs in `data_subset`: the main "full"
# holdout (5,405 points/model) and a separate auxiliary "iiiF_only" run
# (250/model). Restrict to "full" so every model is scored on the same canonical
# holdout; otherwise metrics blend the two runs (5,655/model).
holdout_df <- read_csv(HOLDOUT_EVAL, show_col_types = FALSE) %>%
  filter(data_subset == "full") %>%
  mutate(
    S5          = if_else(insert_5_id == "eNMU_region_iii" &
                          insert_5_ori == "forward",
                          "e-NMU(iii)F", "Other"),
    model_label = paste0(family, "/", model)
  )

cat(nrow(holdout_df), "holdout predictions loaded.\n")

# R² and Pearson r per model
model_r2 <- holdout_df %>%
  group_by(model_label) %>%
  summarise(
    r2         = r2_score(actual_avg_Rep, predicted_avg_Rep),
    pearson_r  = cor(actual_avg_Rep, predicted_avg_Rep, method = "pearson"),
    pearson_r2 = pearson_r^2,
    n          = n(),
    .groups = "drop"
  )

model_order <- model_r2 %>% arrange(desc(r2)) %>% pull(model_label)
best_model  <- model_order[1]
best_rf     <- model_order[2]
cat("Best RF model:", best_rf,
    "| R² =", round(model_r2$r2[model_r2$model_label == best_rf], 3), "\n")
```

## Load downsampling sweep (50 permutations)
`downsampling_metrics.csv` holds one row per (model × permutation × training
fraction).
We collapse across the 50 permutations to a per-fraction **mean** and a **t-based
95% confidence interval** (mean ± t·SE, where SE = SD / √n_perm and t is the
0.975 quantile of Student's t with n_perm − 1 degrees of freedom). The t
interval — rather than the normal-approximation 1.96·SE. 
The `pearson_r` column of the resulting table holds
the mean, so the downstream threshold / smoothing chunks operate on the averaged
curve.

Two parameters control the panels:

- `downsample_stepsize` — thin the 1%-resolution curve to every X% for the
  plotted points/line, anchored at 100% and stepping downward. The CI band and
  the sufficiency threshold are always computed on the full 1% grid.
- `downsample_model` — which model arm to plot (`"rf"` or `"xgb"`); drives the
  curve data and the saved-figure filenames. 

```{r load-downsampling}
downsample_stepsize <- 3
downsample_model    <- "rf"
ci_level            <- 0.95   # confidence level for the CI on the per-fraction mean

# Raw per-permutation sweep for the selected model arm.
downsample_raw <- read_csv(DOWNSAMPLE, show_col_types = FALSE) %>%
  filter(model == downsample_model)

n_perm <- n_distinct(downsample_raw$replicate)
cat(sprintf("%d downsampling permutations × %d fractions for model '%s'.\n",
            n_perm, n_distinct(downsample_raw$fraction), downsample_model))

stopifnot(downsample_model %in% downsample_raw$model)

# Display label for the selected model.
ds_model_label <- c(rf = "Random forest", xgb = "Gradient boosting")[downsample_model]

# ── Collapse across permutations: per-fraction mean + 95% CI ──────────────────
# `pearson_r` is kept as the MEAN across permutations so the threshold /
# smoothing chunks below reuse it unchanged. pct_removed / pct_used give the two
# complementary x-axis framings (fraction of the full training set removed vs.
# retained).
ds_ci_full <- downsample_raw %>%
  group_by(model, fraction, n_train) %>%
  summarise(
    n_rep          = n(),
    # NOTE: compute SD/SE from the raw per-permutation `pearson_r` BEFORE
    # overwriting `pearson_r` with its mean (dplyr summarise resolves columns
    # sequentially, so the mean must be assigned after the SD/SE that need the
    # raw values).
    pearson_r_sd   = sd(pearson_r),
    pearson_r_se   = pearson_r_sd / sqrt(n_rep),
    pearson_r      = mean(pearson_r),            # mean across permutations
    # t-based 95% CI on the mean: critical value from Student's t with
    # (n_rep - 1) df, computed PER FRACTION so it stays correct if any fraction
    # has fewer permutations. At n_rep = 50 this is t = 2.01 (vs. 1.96 normal).
    pearson_r_tcrit = qt(1 - (1 - ci_level) / 2, df = n_rep - 1),
    pearson_r_lo   = pearson_r - pearson_r_tcrit * pearson_r_se,
    pearson_r_hi   = pearson_r + pearson_r_tcrit * pearson_r_se,
    pearson_r2     = mean(pearson_r2),
    r2_coef_determination = mean(r2_coef_determination),
    .groups = "drop"
  ) %>%
  arrange(n_train) %>%
  mutate(pct_used    = fraction * 100,
         pct_removed = 100 - fraction * 100)

# Alias used by the threshold / smoothing chunks (per-fraction averaged curve).
ds_curve_full <- ds_ci_full

# Thinned copy for the plotted points/line (every downsample_stepsize %).
# round() guards against floating-point drift in fraction (e.g. 0.07*100).
ds_ci <- ds_ci_full %>%
  filter((100 - round(fraction * 100)) %% downsample_stepsize == 0)

# Full-data (fraction = 1.0) mean holdout Pearson r — saturation reference.
ds_full_r2 <- ds_ci_full %>%
  filter(fraction == 1.0) %>%
  pull(pearson_r)

cat(ds_model_label, "mean holdout Pearson r: ",
    sprintf("%.3f @ %d train rows (%.0f%%)  →  %.3f @ %d train rows (full)\n",
            ds_ci_full$pearson_r[1],
            ds_ci_full$n_train[1],
            100 * ds_ci_full$fraction[1],
            ds_full_r2,
            max(ds_ci_full$n_train)))
```


# Figure 2
## Fig 2a 
```{r}
id_cols_all  <- paste0("insert_", 1:5, "_id")
id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

line_data_ivrev <- rfxgb_full %>%
  mutate(
    group = case_when(
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "forward" ~ "e-NMU(iii)F",
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "reverse" ~ "e-NMU(iii)R",
      insert_5_id == "eNMU_region_iv"  & insert_5_ori == "reverse" ~ "e-NMU(iv)R",
      TRUE ~ "minP"
    ),
    # iiiF / iiiR / ivR: count enhancers in S1-S4 only (range 0-4)
    # minP: count enhancers in S1-S5 (range 0-5)
    total_enmu = if_else(
      group == "minP",
      rowSums(across(all_of(id_cols_all),  ~ as.integer(str_detect(.x, "eNMU_region")))),
      rowSums(across(all_of(id_cols_s1s4), ~ as.integer(str_detect(.x, "eNMU_region"))))
    )
  )

# ── print total n per line (group) ────────────────────────────────────────────
cat("\n── Total constructs per line ──\n")
print(line_data_ivrev %>% count(group, name = "n_total"))

line_data_ivrev_summary <- line_data_ivrev %>%
  group_by(group, total_enmu) %>%
  summarise(
    mean_rep_log = mean(log10(avg_Rep + 0.05)),
    se_log       = sd(log10(avg_Rep + 0.05)) / sqrt(n()),
    ci95_log     = se_log * qt(0.975, n() - 1),
    n            = n(),
    .groups      = "drop"
  )

p_fig2a_iiirev_ivrev <- ggplot(line_data_ivrev_summary,
                        aes(x = total_enmu, y = 10^mean_rep_log,
                            color = group, shape = group)) +
  geom_errorbar(aes(ymin = 10^(mean_rep_log - ci95_log),
                    ymax = 10^(mean_rep_log + ci95_log)),
                width = 0.2, alpha = 0.6) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8, stroke = 0.8) +
  scale_color_manual(
    values = c(
      "e-NMU(iii)F" = unname(COLOR_IIIF),
      "e-NMU(iii)R" = unname(COLOR_IIIF),
      "e-NMU(iv)R"  = unname(enmu_palette[4]),
      "minP"        = unname(COLOR_OTHER)
    )
  ) +
  scale_shape_manual(
    values = c(
      "e-NMU(iii)F" = 16,
      "e-NMU(iii)R" = 4,
      "e-NMU(iv)R"  = 16,
      "minP"        = 16
    )
  ) +
  scale_y_log10(labels = scales::label_log()) +
  annotation_logticks(sides = "l", size = 0.3) +
  labs(
    x     = "# of enhancers in upstream slots",
    y     = "Activity score",
    color = "Promoter group",
    shape = "Promoter group"
  ) +
  theme(
    axis.title.y     = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
print(p_fig2a_iiirev_ivrev)
save_fig("fig2a_upstream_enh", p_fig2a_iiirev_ivrev, panel_width = PANEL_W, panel_height = PANEL_H)
```

## Fig 2b
```{r}
id_cols_s1s4_only <- paste0("insert_", 1:4, "_id")

elem_copy_map_s1s4 <- c(
  "eNMU_region_i"       = "e-NMU(i)",
  "eNMU_region_ii"      = "e-NMU(ii)",
  "eNMU_region_iii"     = "e-NMU(iii)",
  "eNMU_region_iv"      = "e-NMU(iv)",
  "eNMU_region_v"       = "e-NMU(v)",
  "synthetic_insulator" = "Insu(i)"
)

cat(sprintf("Total constructs in rfxgb_full (n = %s) used for this panel.\n",
            scales::comma(nrow(rfxgb_full))))

# ── per-construct copy_n (raw, un-summarised) for each element, S1–S4 only ──
copy_n_raw_s1s4 <- map_dfr(names(elem_copy_map_s1s4), function(elem_id) {
  rfxgb_full %>%
    mutate(copy_n = rowSums(across(all_of(id_cols_s1s4_only), ~ .x == elem_id))) %>%
    transmute(
      element_label = elem_copy_map_s1s4[[elem_id]],
      copy_n,
      avg_Rep
    )
}) %>%
  mutate(
    # collapse 3 and 4 into "3+" before summarising, so mean/CI are computed
    # on the pooled raw data rather than merged from separate bins
    copy_n_bin = if_else(copy_n >= 3, "3+", as.character(copy_n)),
    copy_n_bin = factor(copy_n_bin, levels = c("0", "1", "2", "3+"))
  )

copy_n_data_s1s4 <- copy_n_raw_s1s4 %>%
  group_by(element_label, copy_n_bin) %>%
  summarise(
    mean_log = mean(log10(avg_Rep + 0.05)),
    se_log   = sd(log10(avg_Rep + 0.05)) / sqrt(n()),
    ci95_log = se_log * qt(0.975, pmax(n() - 1, 1)),
    n        = n(),
    .groups  = "drop"
  )

copy_n_log_s1s4 <- copy_n_data_s1s4 %>%
  mutate(element_label = factor(as.character(element_label),
                                levels = unname(elem_copy_map_s1s4))) %>%
  filter(element_label != "Insu(i)")   # exclude insulator, matching fig2b

# ── console table: n per datapoint on each line ─────────────────────────────
cat("\n── n per datapoint (element × copy_n bin), S1–S4 only, 3+ collapsed ──\n")
print(
  copy_n_log_s1s4 %>%
    arrange(element_label, copy_n_bin) %>%
    select(element_label, copy_n_bin, n),
  n = Inf
)

# ── compute y-axis limits from the actual CI bounds so nothing is clipped ───
y_lo_data <- min(10^(copy_n_log_s1s4$mean_log - copy_n_log_s1s4$ci95_log))
y_hi_data <- max(10^(copy_n_log_s1s4$mean_log + copy_n_log_s1s4$ci95_log))

# a little multiplicative padding so error bars don't sit flush on the edge
y_lo_plot <- min(10^-1, y_lo_data / 1.15)
y_hi_plot <- max(10^1,  y_hi_data * 1.15)

cat(sprintf("\nY-axis range needed to fit all CI bounds: %.3f – %.3f\n",
            y_lo_data, y_hi_data))
cat(sprintf("Y-axis plot limits used (with padding):    %.3f – %.3f\n",
            y_lo_plot, y_hi_plot))

p_copy_log_comb_s1s4 <- ggplot(copy_n_log_s1s4,
                           aes(x = copy_n_bin, y = 10^mean_log,
                               color = element_label, group = element_label)) +
  geom_errorbar(
    aes(ymin = 10^(mean_log - ci95_log), ymax = 10^(mean_log + ci95_log)),
    width = 0.15, alpha = 0.4
  ) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  scale_y_log10(
    breaks = 10^(-1:1),
    labels = scales::label_log(),
    limits = c(y_lo_plot, y_hi_plot)
  ) +
  annotation_logticks(sides = "l", size = 0.3) +
  scale_color_manual(values = elem_colors, name = "Element") +
  labs(
    x = "Element copy number (S1–S4)",
    y = "Activity score"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

print(p_copy_log_comb_s1s4)
save_fig("fig2b_copy_number", p_copy_log_comb_s1s4,
         panel_width = PANEL_W * 1.1, panel_height = PANEL_H)
```

## Fig 2c
```{r}
## Fig 2c′ (combined) — Relative Activity vs. Ctrl Pool by Slot (iiiF-S5 Subset)
## Single panel, three lines: e-NMU(iii), e-NMU(iv), Insu(i) — each pools BOTH
## orientations. e-NMU(iii)/(iv) numerators are restricted to constructs with
## NO insulator anywhere in S1–S4 (removes insulator co-occurrence as a
## confound on those two lines). The Ctrl-pool baseline and the Insu(i) line
## are NOT restricted — they use the original, full iiiF subset.

id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

# ── iiiF subset (self-contained rebuild, matches earlier `df_iiiF`) ─────────
df_iiiF_standalone <- df %>% filter(S5 == "e-NMU(iii)F")
cat(nrow(df_iiiF_standalone), "constructs in iiiF subset.\n")

# ── insulator-free version, for the e-NMU(iii)/(iv) numerators only ────────
df_iiiF_no_insu_standalone <- df_iiiF_standalone %>%
  filter(!if_any(all_of(id_cols_s1s4), ~ .x == "synthetic_insulator"))

cat(nrow(df_iiiF_no_insu_standalone),
    "constructs after removing insulator-bearing constructs (e-NMU(iii)/(iv) only).\n")

# ── long tables: full (for Ctrl + Insu) and insulator-free (for e-NMU) ─────
elements_long_three_full    <- build_elements_long(df_iiiF_standalone)
elements_long_three_no_insu <- build_elements_long(df_iiiF_no_insu_standalone)

elements_log_three_full    <- elements_long_three_full    %>% mutate(log_act = log10(avg_Rep + 0.05))
elements_log_three_no_insu <- elements_long_three_no_insu %>% mutate(log_act = log10(avg_Rep + 0.05))

# ── Ctrl-pool baseline per slot — from the FULL (unrestricted) subset ──────
ctrl_baseline_three <- elements_log_three_full %>%
  filter(family == "Ctrl") %>%
  group_by(slot) %>%
  summarise(
    mean_log_ctrl = mean(log_act),
    se_log_ctrl   = sd(log_act) / sqrt(n()),
    n_ctrl        = n(),
    .groups = "drop"
  )

cat("\n── Ctrl-pool n per slot (unrestricted) ──\n")
print(ctrl_baseline_three %>% select(slot, n_ctrl))

# ── e-NMU(iii)/(iv) element means — from the INSULATOR-FREE subset ─────────
elem_mean_log_enmu <- elements_log_three_no_insu %>%
  filter(element %in% c("e-NMU(iii)", "e-NMU(iv)")) %>%
  group_by(family, element, slot) %>%
  summarise(
    mean_log_elem = mean(log_act),
    se_log_elem   = sd(log_act) / sqrt(n()),
    n_elem        = n(),
    .groups = "drop"
  )

# ── Insu(i) element means — from the FULL (unrestricted) subset ────────────
elem_mean_log_insu <- elements_log_three_full %>%
  filter(element == "Insu(i)") %>%
  group_by(family, element, slot) %>%
  summarise(
    mean_log_elem = mean(log_act),
    se_log_elem   = sd(log_act) / sqrt(n()),
    n_elem        = n(),
    .groups = "drop"
  )

elem_mean_log_three <- bind_rows(elem_mean_log_enmu, elem_mean_log_insu)

cat("\n── n per (element, slot) ──\n")
print(elem_mean_log_three %>% select(element, slot, n_elem))

# ── combine: log2 FC vs. Ctrl pool, same algebra as original chunk ─────────
target_elements <- c("e-NMU(iii)", "e-NMU(iv)", "Insu(i)")

relative_summary_three <- elem_mean_log_three %>%
  inner_join(ctrl_baseline_three, by = "slot") %>%
  mutate(
    delta_log  = mean_log_elem - mean_log_ctrl,
    se_delta   = sqrt(se_log_elem^2 + se_log_ctrl^2),
    ci95_delta = se_delta * 1.96,
    log2_fc    = delta_log / log10(2),
    log2_lo    = (delta_log - ci95_delta) / log10(2),
    log2_hi    = (delta_log + ci95_delta) / log10(2),
    element    = factor(as.character(element), levels = target_elements)
  )

# ── single-panel plot, three lines, color legend titled "Element" ──────────
p_fig2c_three_combined <- ggplot(relative_summary_three,
                  aes(x = slot, y = log2_fc,
                      color = element, group = element)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  geom_errorbar(aes(ymin = log2_lo, ymax = log2_hi),
                width = 0.15, alpha = 0.6) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  scale_color_manual(values = elem_colors_rel, name = "Element") +
  scale_x_continuous(breaks = 1:4, labels = paste0("S", 1:4)) +
  coord_cartesian(ylim = c(-2, 2.5)) +
  labs(
    x     = "Slot position",
    y     = expression(log[2] ~ "(FC vs. Ctrl pool)")
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

print(p_fig2c_three_combined)
save_fig("fig2c_relative_activity_iii_iv_insu",
         p_fig2c_three_combined, panel_width = PANEL_W, panel_height = PANEL_H)


## ── unique sCRLs contributing to each numerator line (not the Ctrl denominator) ──

id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

n_enmu_iii <- df_iiiF_no_insu_standalone %>%
  filter(if_any(all_of(id_cols_s1s4), ~ .x == "eNMU_region_iii")) %>%
  nrow()

n_enmu_iv <- df_iiiF_no_insu_standalone %>%
  filter(if_any(all_of(id_cols_s1s4), ~ .x == "eNMU_region_iv")) %>%
  nrow()

n_insu <- df_iiiF_standalone %>%
  filter(if_any(all_of(id_cols_s1s4), ~ .x == "synthetic_insulator")) %>%
  nrow()

cat(sprintf("e-NMU(iii) numerator: n = %d unique sCRLs (insulator-free subset)\n", n_enmu_iii))
cat(sprintf("e-NMU(iv) numerator:  n = %d unique sCRLs (insulator-free subset)\n", n_enmu_iv))
cat(sprintf("Insu(i) numerator:    n = %d unique sCRLs (full subset)\n", n_insu))
```

## Fig 2d and Fig 2e
```{r}
id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

# ── Filter: iiiF-S5, exactly one eNMU(iii) or eNMU(iv) in S1–S4,
#            exactly one insulator in S1–S4, no other eNMU anywhere in S1–S4
df_base <- df %>%
  filter(S5 == "e-NMU(iii)F") %>%
  rowwise() %>%
  mutate(
    n_insu       = sum(c_across(all_of(id_cols_s1s4)) == "synthetic_insulator"),
    n_enmu_iiiiv = sum(str_detect(c_across(all_of(id_cols_s1s4)), "eNMU_region_(iii|iv)")),
    n_enmu_any   = sum(str_detect(c_across(all_of(id_cols_s1s4)), "eNMU_region"))
  ) %>%
  ungroup() %>%
  filter(n_insu == 1, n_enmu_iiiiv == 1, n_enmu_any == 1) %>%
  rowwise() %>%
  mutate(
    insu_slot = which(c_across(all_of(id_cols_s1s4)) == "synthetic_insulator"),
    enmu_slot = which(str_detect(c_across(all_of(id_cols_s1s4)), "eNMU_region_(iii|iv)"))
  ) %>%
  ungroup() %>%
  filter(insu_slot != enmu_slot) %>%
  mutate(
    enmu_side = if_else(enmu_slot < insu_slot, "eNMU left of insulator",
                                               "eNMU right of insulator"),
    enmu_side = factor(enmu_side, levels = c("eNMU right of insulator",
                                             "eNMU left of insulator"))
  )

cat(nrow(df_base), "constructs after filtering.\n")

# ── Grouped bar: left vs right ────────────────────────────────────────────────
df_grouped_summary <- df_base %>%
  mutate(log_act = log10(avg_Rep + 0.05)) %>%
  group_by(enmu_side) %>%
  summarise(
    mean_log = mean(log_act),
    se_log   = sd(log_act) / sqrt(n()),
    ci95_log = se_log * qt(0.975, pmax(n() - 1, 1)),
    n        = n(),
    .groups  = "drop"
  ) %>%
  mutate(
    label = factor(
      paste0(enmu_side, " (n = ", n, ")"),
      levels = paste0(levels(enmu_side), " (n = ",
                      n[match(levels(enmu_side), enmu_side)], ")")
    )
  )

cat("\n── Grouped summary ──\n")
print(df_grouped_summary)

p_grouped <- ggplot(df_grouped_summary,
                    aes(x = 10^mean_log, y = label)) +
  geom_col(fill = unname(COLOR_OTHER), alpha = 0.85, width = 0.65) +
  geom_errorbarh(
    aes(xmin = 10^(mean_log - ci95_log),
        xmax = 10^(mean_log + ci95_log)),
    height = 0.25, linewidth = 0.5, color = "grey30"
  ) +
  scale_x_log10(
    breaks = c(1, 10),
    labels = scales::label_log(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  annotation_logticks(sides = "b", size = 0.3) +
  labs(x = "Activity score", y = NULL) +
  theme(
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(family = FONT, size = 9)
  )

print(p_grouped)
save_fig("fig2d_insu_lr_grouped", p_grouped, panel_width = PANEL_W * 0.8, panel_height = PANEL_H)

# ── Individual bar: every (insu_slot × enmu_slot) combination ────────────────
df_indiv_summary <- df_base %>%
  mutate(log_act = log10(avg_Rep + 0.05)) %>%
  group_by(enmu_side, insu_slot, enmu_slot) %>%
  summarise(
    mean_log = mean(log_act),
    se_log   = sd(log_act) / sqrt(n()),
    ci95_log = se_log * qt(0.975, pmax(n() - 1, 1)),
    n        = n(),
    .groups  = "drop"
  ) %>%
  mutate(
    label = pmap_chr(
      list(insu_slot, enmu_slot, n),
      function(ins, enm, nn) {
        slot_contents <- sapply(1:4, function(s) {
          if (s == ins) "Insu"
          else if (s == enm) "eNMU"
          else "·"
        })
        sprintf("[%s]      (n = %d)", paste(slot_contents, collapse = " "), nn)
      }
    ),
    sort_key = if_else(enmu_slot < insu_slot, 0, 1) * 100 + insu_slot * 10 + enmu_slot,
    label    = fct_reorder(label, -sort_key)
  )

cat("\n── Individual summary ──\n")
print(df_indiv_summary %>% select(enmu_side, insu_slot, enmu_slot, n, label))
cat("\nGrouped n check — should match grouped summary above:\n")
print(df_indiv_summary %>% group_by(enmu_side) %>% summarise(total_n = sum(n)))

p_indiv <- ggplot(df_indiv_summary,
                  aes(x = 10^mean_log, y = label)) +
  geom_col(fill = unname(COLOR_OTHER), alpha = 0.85, width = 0.65) +
  geom_errorbarh(
    aes(xmin = 10^(mean_log - ci95_log),
        xmax = 10^(mean_log + ci95_log)),
    height = 0.25, linewidth = 0.5, color = "grey30"
  ) +
  scale_x_log10(labels = scales::label_log(),
                expand  = expansion(mult = c(0, 0.12))) +
  annotation_logticks(sides = "b", size = 0.3) +
  labs(x = "Activity score", y = NULL) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(family = FONT, size = 9)
  )

print(p_indiv)
save_fig("fig2e_insu_lr_individual_bars", p_indiv, panel_width = PANEL_W *2.2, panel_height = PANEL_H *1.3)

cat("\n── Fold change: eNMU right of insulator / eNMU left of insulator ──\n")
fc <- with(df_grouped_summary,
           10^mean_log[enmu_side == "eNMU right of insulator"] /
           10^mean_log[enmu_side == "eNMU left of insulator"])
cat(sprintf("Fold change (right / left): %.2fx\n", fc))
cat(sprintf("log2 FC:                    %.3f\n",  log2(fc)))
```

## Fig 2f
```{r holdout-bar, fig.width=6, fig.height=4}
# Human-readable display names for each raw "family/model" label.
PRETTY_MODEL <- c(
  "rfxgb/xgb"                      = "Gradient boosting",
  "rfxgb/rf"                       = "Random forest",
  "linear-interactions/ols"        = "OLS + int",
  "linear-interactions/ridge"      = "Ridge + int",
  "linear-interactions/lasso"      = "Lasso + int",
  "linear-interactions/elasticnet" = "Elastic net + int",
  "linear/ols"                     = "OLS",
  "linear/ridge"                   = "Ridge",
  "linear/lasso"                   = "Lasso",
  "linear/elasticnet"              = "Elastic net"
)

# Flag any label missing from the map rather than silently keeping the raw form.
unmapped <- setdiff(model_r2$model_label, names(PRETTY_MODEL))
if (length(unmapped)) warning("Unmapped model labels: ",
                              paste(unmapped, collapse = ", "))

# ── Pearson r ─────────────────────────────────────────────────────────────────
# Recode to pretty names, THEN order by the metric so the reorder acts on the
# display labels. .desc = TRUE puts the largest value in the first factor level,
# which ggplot renders at the BOTTOM of the discrete y-axis → best model at bottom.
model_r_bar <- model_r2 %>%
  mutate(
    model_pretty = recode(model_label, !!!PRETTY_MODEL),
    model_pretty = fct_reorder(model_pretty, pearson_r, .desc = TRUE)
  )

p_fig2f_bar <- ggplot(model_r_bar, aes(x = pearson_r, y = model_pretty)) +
  geom_col(fill = "gray", alpha = 1, width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", pearson_r)),
            hjust = -0.1, size = ANNOT_SIZE, family = FONT) +
  scale_x_continuous(limits = c(0, 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = expression(Test ~ set ~ italic(r)), y = NULL) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

print(p_fig2f_bar)
save_fig("fig2f_bar_pearsonr_by_model", p_fig2f_bar, panel_width = PANEL_W * 0.9, panel_height = PANEL_H)
```



```{r per-context-metrics}

# Filter the per-construct holdout predictions to iiiF, ivR, iiiR-at-S5, then summarize.
# Restrict to data_subset == "full": the combined file also stacks a separate
# "iiiF_only" evaluation run, which would otherwise add 8 extra points/model
# (233 full + 8 iiiF_only = 241).
holdout_promoters_only <- holdout_df %>%
  filter(data_subset == "full") %>%
  filter(insert_5_id == "eNMU_region_iii" | insert_5_id == "eNMU_region_iv") %>%
  filter(!(insert_5_id == "eNMU_region_iv" & insert_5_ori == "forward")) #not iiiF

cat(nrow(holdout_promoters_only), "promoter-S5 holdout points",
    sprintf("(%d per model).\n", nrow(holdout_promoters_only) / n_distinct(holdout_promoters_only$model_label)))
# 718 per model

holdout_except_promoters <- holdout_df %>%
  filter(data_subset == "full") %>%
  filter(insert_5_id != "eNMU_region_iii") %>% #not iiiF, iiiR
  filter(!(insert_5_id == "eNMU_region_iv" & insert_5_ori == "reverse")) #not ivR

cat(nrow(holdout_except_promoters), "not-promoter-S5 holdout points",
    sprintf("(%d per model).\n", nrow(holdout_except_promoters) / n_distinct(holdout_except_promoters$model_label)))
# 4687 per model

# 4687 + 718 = 5405 -- good


holdout_iiiF <- holdout_df %>%
  filter(data_subset == "full",
         insert_5_id == "eNMU_region_iii", insert_5_ori == "forward")

holdout_iiiR <- holdout_df %>%
  filter(data_subset == "full",
         insert_5_id == "eNMU_region_iii", insert_5_ori == "reverse")

holdout_ivR <- holdout_df %>%
  filter(data_subset == "full",
         insert_5_id == "eNMU_region_iv", insert_5_ori == "reverse")

cat(nrow(holdout_iiiF), "iiiF-S5 holdout points",
    sprintf("(%d per model).\n", nrow(holdout_iiiF) / n_distinct(holdout_iiiF$model_label)))
cat(nrow(holdout_iiiR), "iiiR-S5 holdout points",
    sprintf("(%d per model).\n", nrow(holdout_iiiR) / n_distinct(holdout_iiiR$model_label)))
cat(nrow(holdout_ivR), "ivR-S5 holdout points",
    sprintf("(%d per model).\n", nrow(holdout_ivR) / n_distinct(holdout_ivR$model_label)))

# 233 + 151 + 334 = 718, good

### MAIN TEXT r for iiiF only
model_r2_iiiF <- holdout_iiiF %>%
  group_by(model_label) %>%
  summarise(
    r2         = r2_score(actual_avg_Rep, predicted_avg_Rep),
    pearson_r  = cor(actual_avg_Rep, predicted_avg_Rep, method = "pearson"),
    pearson_r2 = pearson_r^2,
    n          = n(),
    .groups = "drop"
  )
model_r2_iiiR <- holdout_iiiR %>%
  group_by(model_label) %>%
  summarise(
    r2         = r2_score(actual_avg_Rep, predicted_avg_Rep),
    pearson_r  = cor(actual_avg_Rep, predicted_avg_Rep, method = "pearson"),
    pearson_r2 = pearson_r^2,
    n          = n(),
    .groups = "drop"
  )
model_r2_ivR <- holdout_ivR %>%
  group_by(model_label) %>%
  summarise(
    r2         = r2_score(actual_avg_Rep, predicted_avg_Rep),
    pearson_r  = cor(actual_avg_Rep, predicted_avg_Rep, method = "pearson"),
    pearson_r2 = pearson_r^2,
    n          = n(),
    .groups = "drop"
  )

### r for promoters only, grouped
model_r2_promoters_only <- holdout_promoters_only %>%
  group_by(model_label) %>%
  summarise(
    r2         = r2_score(actual_avg_Rep, predicted_avg_Rep),
    pearson_r  = cor(actual_avg_Rep, predicted_avg_Rep, method = "pearson"),
    pearson_r2 = pearson_r^2,
    n          = n(),
    .groups = "drop"
  )


### MAIN TEXT r for except-promoters
model_r2_except_promoters <- holdout_except_promoters %>%
  group_by(model_label) %>%
  summarise(
    r2         = r2_score(actual_avg_Rep, predicted_avg_Rep),
    pearson_r  = cor(actual_avg_Rep, predicted_avg_Rep, method = "pearson"),
    pearson_r2 = pearson_r^2,
    n          = n(),
    .groups = "drop"
  )
```

## Fig 2g
```{r}
# ── Select the rfxgb/rf model specifically ───────────────────────────────────
target_model <- "rfxgb/rf"

if (!target_model %in% model_r2$model_label) {
  stop(sprintf("model_label '%s' not found. Available labels: %s",
               target_model, paste(model_r2$model_label, collapse = ", ")))
}

cat("Selected model:", target_model,
    "| Pearson r =", round(model_r2$pearson_r[model_r2$model_label == target_model], 3), "\n")

holdout_best <- holdout_df %>% filter(model_label == target_model)

r_best <- cor(holdout_best$actual_avg_Rep,
              holdout_best$predicted_avg_Rep,
              method = "pearson")

# ── Rebuild S5 fresh, collapsing everything non-target into "All other" ─────
holdout_best <- holdout_best %>%
  mutate(
    S5 = case_when(
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "forward" ~ "e-NMU(iii)F",
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "reverse" ~ "e-NMU(iii)R",
      insert_5_id == "eNMU_region_iv"  & insert_5_ori == "reverse" ~ "e-NMU(iv)R",
      TRUE                                                        ~ "All other"
    ),
    S5 = factor(S5, levels = c("e-NMU(iii)F", "e-NMU(iii)R", "e-NMU(iv)R", "All other"))
  )

set.seed(42)
best_sample <- holdout_best %>%
  slice_sample(n = min(20000, nrow(.))) %>%
  mutate(
    actual_plot    = pmax(actual_avg_Rep    + 0.05, 0.05),
    predicted_plot = pmax(predicted_avg_Rep + 0.05, 0.05)
  )

cat(sprintf("Model: %s\n", target_model))
cat(sprintf("Total points plotted: %d\n", nrow(best_sample)))
print(table(best_sample$S5))

set.seed(42)
best_shuffled <- best_sample %>% slice_sample(prop = 1)

r_label <- sprintf('italic(r) ~ "= %.2f"', r_best)

base_scales <- list(
  scale_x_log10(labels = scales::label_log()),
  scale_y_log10(labels = scales::label_log()),
  annotation_logticks(sides = "bl", size = 0.3),
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black"),
  labs(x = "Actual activity score", y = "Predicted activity score"),
  annotate("text", x = 1.04, y = Inf, hjust = 0, vjust = 1.5,
           label = r_label, parse = TRUE, size = ANNOT_SIZE),
  theme(
    legend.position      = "bottom",
    legend.justification = "center",
    panel.grid.major     = element_blank(),
    panel.grid.minor     = element_blank()
  )
)

p_fig2g <- ggplot(best_shuffled,
       aes(x = actual_plot, y = predicted_plot, color = S5,
           shape = S5, size = S5, alpha = S5, stroke = S5)) +
  geom_point() +
  scale_color_manual(
    values = c(
      "All other"   = unname(COLOR_OTHER),
      "e-NMU(iv)R"  = unname(enmu_palette[4]),
      "e-NMU(iii)F" = unname(COLOR_IIIF),
      "e-NMU(iii)R" = unname(COLOR_IIIF)
    ),
    name = "S5 identity"
  ) +
  scale_shape_manual(
    values = c("All other" = 16, "e-NMU(iv)R" = 16,
               "e-NMU(iii)F" = 16, "e-NMU(iii)R" = 4),
    name = "S5 identity"
  ) +
  scale_size_manual(
    values = c("All other" = 0.5, "e-NMU(iv)R" = 0.5,
               "e-NMU(iii)F" = 0.5, "e-NMU(iii)R" = 0.5),
    name = "S5 identity"
  ) +
  scale_alpha_manual(
    values = c("All other" = 0.7, "e-NMU(iv)R" = 0.7,
               "e-NMU(iii)F" = 0.7, "e-NMU(iii)R" = 0.7),
    name = "S5 identity"
  ) +
  scale_discrete_manual("stroke",
    values = c("All other" = 0.3, "e-NMU(iv)R" = 0.3,
               "e-NMU(iii)F" = 0.3, "e-NMU(iii)R" = 0.3),
    name = "S5 identity"
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 1.5), nrow = 2),
    shape = guide_legend(nrow = 2),
    alpha = guide_legend(nrow = 2),
    size  = guide_legend(nrow = 2)
  ) +
  base_scales

print(p_fig2g)
save_fig("fig2g_rf_scatter", p_fig2g, panel_width = PANEL_W * 0.9)
```

## Fig 2h — Holdout Learning Curve (mean ± 95% CI across 50 permutations)
Holdout Pearson r of the selected model (`downsample_model`, currently
**`r ds_model_label`**) as a function of the **% of training data removed**. 
The solid line is the **mean across the 50 downsampling permutations** and the 
shaded band is the **t-based 95% confidence interval** 
(mean ± t·SE, t from Student's t with n_perm − 1 df). The
grey box marks the region **within 5% of the full-data (100%) mean
performance** (the curve is plateaued at the top, so "maximum" is anchored to
the 100%-data mean rather than the noise-sensitive empirical argmax). The dotted
line marks the least training data at which the **lower 95% CI bound** still
clears the 95%-of-max threshold.
```{r fig2h-learning-curve-pearsonr-ci, fig.width=4.3, fig.height=3.3}
# Max ±5% band. "Maximum" is anchored to the FULL-DATA (fraction == 1.0) mean
# rather than the empirical argmax of the curve: the curve is plateaued across
# the top few percent, so the argmax floats to a fraction < 100% purely by
# random-forest / subsampling noise (99% beats 100% by ~2e-5). Anchoring to the
# 100%-data mean makes "maximum" = "performance with all the data", matching the
# plain reading of the sufficiency claim and removing the noise sensitivity.
pr_max  <- ds_ci_full$pearson_r[ds_ci_full$fraction == 1.0]
band_lo <- 0.95 * pr_max
band_hi <- 1.05 * pr_max

# Data-sufficiency threshold: the smallest training fraction at which the lower
# 95% CI bound of the mean Pearson r still clears 95% of the maximum.
thresh_row <- ds_ci_full %>%
  filter(pearson_r_lo >= band_lo) %>%
  slice_min(fraction, n = 1)

pct_used_thresh    <- thresh_row$pct_used      # % of full training data that suffices
pct_removed_thresh <- thresh_row$pct_removed   # complementary % removed

cat(sprintf("mean Pearson r max = %.4f | 95%%-of-max = %.4f\n", pr_max, band_lo))
cat(sprintf("lower 95%% CI clears 95%%-of-max at %.0f%% of data used (%.0f%% removed), n_train = %d (mean r = %.4f, lo = %.4f)\n",
            pct_used_thresh, pct_removed_thresh, thresh_row$n_train,
            thresh_row$pearson_r, thresh_row$pearson_r_lo))

p_fig2h_pearsonr_ci <- ggplot(ds_ci_full, aes(x = pct_removed, y = pearson_r)) +
  # Grey box: within 5% of maximum mean performance.
  annotate("rect", xmin = 0, xmax = 100, ymin = band_lo, ymax = band_hi,
           fill = "grey80", alpha = 0.6) +
  annotate("text", x = 100, y = band_hi,
           label = "Max~italic(r) ~ '± 5%'", parse = TRUE,
           hjust = 1, vjust = 4.5, size = 3, family = FONT, color = "grey40") +
  # Data-sufficiency threshold (lower-CI crossing), in "% removed" space.
  geom_vline(xintercept = pct_removed_thresh, linetype = "dotted",
             color = "grey30", linewidth = 0.5) +
  # 95% CI band across permutations (full 1% grid).
  ## UN/COMMENT THE BELOW TO TOGGLE 95% CI BAND
  geom_ribbon(aes(ymin = pearson_r_lo, ymax = pearson_r_hi),
              fill = unname(COLOR_LINE), alpha = 0.18) +
  geom_line(linewidth = 0.5, color = unname(COLOR_LINE)) +
  # # Thinned points to declutter the mean curve.
  # geom_point(data = ds_ci, size = 0.8, color = unname(COLOR_LINE)) +
  scale_x_continuous(labels = function(x) paste0(x, "%"),
                     breaks = seq(0, 100, 25), limits = c(0, 100)) +
  labs(
    x = "Training data removed (%)",
    y = expression(Test ~ set ~ italic(r))
  ) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

print(p_fig2h_pearsonr_ci)
save_fig(paste0("fig2h_", downsample_model, "_learning_curve_pearsonr_ci"),
         p_fig2h_pearsonr_ci, panel_width = PANEL_W * 0.9, panel_height = PANEL_H)
```

### Main-text sentence

We next performed a downsampling analysis in which we systematically reduced the
training set size and found that **`r sprintf("%.0f%%", pct_used_thresh)`** of
the full training data was sufficient to recover 95% of the maximum Pearson r
(Fig. 2h), ...

### Exact 95%-of-max threshold on the mean curve (1% resolution)

For reference, the crossing computed on the **mean** curve alone (no CI), at the
full 1% resolution.

```{r fig2h-threshold-1pct-resolution}
cat(sprintf("Full-resolution sweep: %d fractions (every %.0f%%) for model '%s'.\n",
            nrow(ds_curve_full),
            100 * min(diff(sort(unique(ds_curve_full$fraction)))),
            downsample_model))

# ── Pearson r (mean curve) ────────────────────────────────────────────────────
# "Maximum" anchored to the full-data (fraction == 1.0) mean (see the CI chunk).
pr_max_full  <- ds_curve_full$pearson_r[ds_curve_full$fraction == 1.0]
band_lo_full <- 0.95 * pr_max_full
row_r <- ds_curve_full %>%
  filter(pearson_r >= band_lo_full) %>%
  slice_min(fraction, n = 1)

cat(sprintf("mean Pearson r max = %.4f | 95%%-of-max = %.4f reached at %.0f%% of data used (n_train = %d, r = %.4f)\n",
            pr_max_full, band_lo_full, row_r$fraction * 100, row_r$n_train, row_r$pearson_r))
```

# Supplementary Figure 2
## Fig S2a
```{r}
id_cols_all  <- paste0("insert_", 1:5, "_id")
id_cols_s1s4 <- paste0("insert_", 1:4, "_id")
ori_cols_all <- paste0("insert_", 1:5, "_ori")

line_data_ivrev <- rfxgb_full %>%
  mutate(
    group = case_when(
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "forward" ~ "e-NMU(iii)F",
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "reverse" ~ "e-NMU(iii)R",
      insert_5_id == "eNMU_region_iv"  & insert_5_ori == "reverse" ~ "e-NMU(iv)R",
      TRUE ~ "minP"
    ),
    # iiiF / iiiR / ivR: count enhancers in S1-S4 only (range 0-4)
    # minP: count enhancers in S1-S5 (range 0-5)
    total_enmu = if_else(
      group == "minP",
      rowSums(across(all_of(id_cols_all),  ~ as.integer(str_detect(.x, "eNMU_region")))),
      rowSums(across(all_of(id_cols_s1s4), ~ as.integer(str_detect(.x, "eNMU_region"))))
    ),
    group = factor(group, levels = c("e-NMU(iii)F", "e-NMU(iii)R", "e-NMU(iv)R", "minP"))
  )

# ── print total n per line (group) ────────────────────────────────────────────
cat("\n── Total constructs per line ──\n")
print(line_data_ivrev %>% count(group, name = "n_total"))

# ── facet title map, matching FigS2a nomenclature: "<name>-S5 (n = ###)" ────
panel_title_map_2a <- c(
  "e-NMU(iii)F" = "e-NMU(iii)F-S5",
  "e-NMU(iii)R" = "e-NMU(iii)R-S5",
  "e-NMU(iv)R"  = "e-NMU(iv)R-S5",
  "minP"        = "minP"
)

panel_n_2a <- line_data_ivrev %>% count(group, name = "n_panel")

panel_labels_2a <- setNames(
  paste0(panel_title_map_2a[levels(line_data_ivrev$group)],
         " (n = ", scales::comma(panel_n_2a$n_panel[match(levels(line_data_ivrev$group), panel_n_2a$group)]), ")"),
  levels(line_data_ivrev$group)
)

line_data_ivrev <- line_data_ivrev %>%
  mutate(panel_label = factor(panel_labels_2a[as.character(group)], levels = panel_labels_2a))

group_colors_2a <- c(
  "e-NMU(iii)F" = unname(COLOR_IIIF),
  "e-NMU(iii)R" = unname(COLOR_IIIF),
  "e-NMU(iv)R"  = unname(enmu_palette[4]),
  "minP"        = unname(COLOR_OTHER)
)
# recolor by panel_label instead of group, since facet now uses panel_label
panel_colors_2a <- setNames(group_colors_2a[levels(line_data_ivrev$group)], panel_labels_2a)

p_fig2a_iiirev_ivrev_violin_facet <- ggplot(line_data_ivrev,
                        aes(x = factor(total_enmu), y = avg_Rep + 0.05,
                            fill = panel_label)) +
  geom_violin(trim = FALSE, alpha = 0.7, color = "grey30", linewidth = 0.3) +
  geom_boxplot(width = 0.1, outlier.shape = NA,
              color = "black", linewidth = 0.3, alpha = 1) +
  scale_fill_manual(values = panel_colors_2a, guide = "none") +
  scale_y_log10(labels = scales::label_log(),
               breaks = 10^(-1:4),
               expand = expansion(mult = c(0.05, 0.25))) +
  annotation_logticks(sides = "l", size = 0.3) +
  facet_wrap(~ panel_label, nrow = 1, scales = "free_x") +
  labs(
    x = "# of enhancers in upstream slots",
    y = "Activity score"
  ) +
  theme(
    axis.title.y     = element_text(hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 8),
    legend.position  = "none"
  )

print(p_fig2a_iiirev_ivrev_violin_facet)
save_fig("figS2a_upstream_enh_violin", p_fig2a_iiirev_ivrev_violin_facet,
         panel_width = PANEL_W * 0.85, panel_height = PANEL_H*0.9)
```

## Fig S2b
```{r}
id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

active_ids   <- c("eNMU_region_iii", "eNMU_region_iv")
inactive_ids <- c("eNMU_region_i", "eNMU_region_ii", "eNMU_region_v")
all_enmu_ids <- c(active_ids, inactive_ids)

# ── classify S5, count eNMU copies, and flag presence of each subset ────────
df_base <- rfxgb_full %>%
  mutate(
    S5_group = case_when(
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "forward" ~ "e-NMU(iii)F",
      insert_5_id == "eNMU_region_iii" & insert_5_ori == "reverse" ~ "e-NMU(iii)R",
      insert_5_id == "eNMU_region_iv"  & insert_5_ori == "reverse" ~ "e-NMU(iv)R",
      TRUE ~ "Other"
    ),
    n_active     = rowSums(across(all_of(id_cols_s1s4), ~ .x %in% active_ids)),
    n_inactive   = rowSums(across(all_of(id_cols_s1s4), ~ .x %in% inactive_ids)),
    has_active   = n_active   > 0,
    has_inactive = n_inactive > 0,
    is_mixed     = has_active & has_inactive
  )

s5_order <- c("e-NMU(iii)F", "e-NMU(iii)R", "e-NMU(iv)R", "Other")

panel_title_map <- c(
  "e-NMU(iii)F" = "e-NMU(iii)F-S5",
  "e-NMU(iii)R" = "e-NMU(iii)R-S5",
  "e-NMU(iv)R"  = "e-NMU(iv)R-S5",
  "Other"       = "minP"
)

# ── UNIQUE n per panel (mixed constructs excluded, counted once) ────────────
panel_n_unique <- df_base %>%
  filter(!is_mixed) %>%
  count(S5_group, name = "n_panel") %>%
  complete(S5_group = s5_order, fill = list(n_panel = 0))

panel_labels <- setNames(
  paste0(panel_title_map[s5_order],
         " (n = ", scales::comma(panel_n_unique$n_panel[match(s5_order, panel_n_unique$S5_group)]), ")"),
  s5_order
)

cat("── Unique n per panel (mixed constructs excluded, counted once) ──\n")
print(panel_n_unique)

# ── build two "pure" long datasets, one per line type ────────────────────────
data_active <- df_base %>%
  filter(!has_inactive) %>%
  transmute(S5_group, avg_Rep,
            line_type = "Active: e-NMU(iii), (iv)",
            copy_n    = n_active)

data_inactive <- df_base %>%
  filter(!has_active) %>%
  transmute(S5_group, avg_Rep,
            line_type = "Inactive: e-NMU(i), (ii), (v)",
            copy_n    = n_inactive)

df_active_inactive_long <- bind_rows(data_active, data_inactive) %>%
  mutate(
    S5_group    = factor(S5_group, levels = s5_order),
    panel_label = factor(panel_labels[as.character(S5_group)], levels = panel_labels),
    line_type   = factor(line_type, levels = c("Active: e-NMU(iii), (iv)",
                                                "Inactive: e-NMU(i), (ii), (v)")),
    # ── collapse copy_n 3 and 4 into "3+" for all panels ───────────────────
    copy_n_bin  = if_else(copy_n >= 3, "3+", as.character(copy_n)),
    copy_n_bin  = factor(copy_n_bin, levels = c("0", "1", "2", "3+"))
  )

cat("\n── Purity filtering: constructs excluded per line ──\n")
cat(sprintf("Active line   — excluded %s constructs with ≥1 inactive element present\n",
            scales::comma(sum(df_base$has_inactive))))
cat(sprintf("Inactive line — excluded %s constructs with ≥1 active element present\n",
            scales::comma(sum(df_base$has_active))))

# ── summarise on the collapsed bin (mean/CI recomputed from raw data) ───────
summary_active_inactive <- df_active_inactive_long %>%
  group_by(panel_label, line_type, copy_n_bin) %>%
  summarise(
    mean_log = mean(log10(avg_Rep + 0.05)),
    se_log   = sd(log10(avg_Rep + 0.05)) / sqrt(n()),
    ci95_log = se_log * qt(0.975, pmax(n() - 1, 1)),
    n        = n(),
    .groups  = "drop"
  )

cat("\n── Bins sorted by smallest n (informational only — nothing dropped) ──\n")
print(summary_active_inactive %>% arrange(n) %>%
        select(panel_label, line_type, copy_n_bin, n, mean_log, ci95_log) %>% head(10))

# ── plot ──────────────────────────────────────────────────────────────────────
Y_MIN <- 0.1   # 10^-1
Y_MAX <- 200   # 10^2

p_active_inactive <- ggplot(summary_active_inactive,
                            aes(x = copy_n_bin, y = 10^mean_log,
                                color = line_type, group = line_type)) +
  geom_errorbar(aes(ymin = 10^(mean_log - ci95_log), ymax = 10^(mean_log + ci95_log)),
                width = 0.15, alpha = 0.6) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  scale_color_manual(
    values = c(
      "Active: e-NMU(iii), (iv)"      = unname(COLOR_IIIF),
      "Inactive: e-NMU(i), (ii), (v)" = unname(COLOR_OTHER)
    ),
    name = "e-NMU elements"
  ) +
  scale_y_log10(labels = scales::label_log()) +
  coord_cartesian(ylim = c(Y_MIN, Y_MAX)) +
  annotation_logticks(sides = "l", size = 0.3) +
  facet_wrap(~ panel_label, nrow = 1, scales = "fixed") +
  labs(
    x = "Element copy number (S1–S4)",
    y = "Activity score"
  ) +
  theme(
    axis.title.y      = element_text(hjust = 0.5),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank(),
    strip.text        = element_text(size = 8)
  )

print(p_active_inactive)
save_fig("figS2b_active_inactive_enmu_by_s5_fourpanel", p_active_inactive,
         panel_width = PANEL_W * 0.68, panel_height = PANEL_H * 0.9)
```

## Fig S2c
```{r figs2c, fig.width=9, fig.height=3.5}
# Factor level order places e-NMU(i–v) on row 1, Ctrl(i–v) on row 2.
# Insulator excluded entirely.
fc_levels_enmu_top <- c(
  "e-NMU(i)", "e-NMU(ii)", "e-NMU(iii)", "e-NMU(iv)", "e-NMU(v)",
  "Ctrl(i)", "Ctrl(ii)", "Ctrl(iii)", "Ctrl(iv)", "Ctrl(v)"
)

ori_pos_fc_enmu_top <- ori_long %>%
  group_by(element_id, element_label, slot, orientation) %>%
  summarise(mean_raw = mean(avg_Rep), n = n(), .groups = "drop") %>%
  pivot_wider(names_from = orientation, values_from = c(mean_raw, n)) %>%
  mutate(
    log2_fc       = log2((mean_raw_forward + 0.05) / (mean_raw_reverse + 0.05)),
  color         = if_else(str_detect(element_id, "eNMU_region"),
                        unname(COLOR_ENMU_GENERIC),
                        get_part_color(paste0("insert_1_id_", element_id))),
    element_label = factor(as.character(element_label), levels = fc_levels_enmu_top)
  ) %>%
  filter(!is.na(element_label))   # drops Insu(i) (not in levels vector)

p_fig2d_fc_enmu_top <- ggplot(ori_pos_fc_enmu_top,
                               aes(x = factor(slot), y = log2_fc, fill = color)) +
  geom_col(show.legend = FALSE) +
  scale_fill_identity() +
  scale_x_discrete(labels = paste0("S", 1:5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#999") +
  facet_wrap(~ element_label, nrow = 2, scales = "fixed") +
  labs(
    x = "Slot position",
    y = expression(log[2] ~ "(Forward / Reverse)")
  ) +
  theme(strip.text = element_text(size = 8))

print(p_fig2d_fc_enmu_top)
save_fig("figS2c_orientation_log2fc", p_fig2d_fc_enmu_top, panel_width = PANEL_W*0.68, panel_height = PANEL_H*0.5)
```

## Fig S2d
```{r}

# ── rebuild ori_long standalone (self-contained; matches the shared build) ──
ori_long_scatter <- map_dfr(1:5, function(s) {
  df %>%
    transmute(
      slot        = s,
      element_id  = .data[[paste0("insert_", s, "_id")]],
      orientation = .data[[paste0("insert_", s, "_ori")]],
      avg_Rep     = avg_Rep
    )
}) %>%
  filter(!is.na(avg_Rep)) %>%
  mutate(
    element_label = element_id %>%
      str_replace_all("eNMU_region_([ivx]+)",           "e-NMU(\\1)") %>%
      str_replace_all("dinucleotide_shuffled_([ivx]+)", "Ctrl(\\1)")   %>%
      str_replace_all("synthetic_insulator",             "Insu(i)")
  )

fc_levels_enmu_top <- c(
  "e-NMU(i)", "e-NMU(ii)", "e-NMU(iii)", "e-NMU(iv)", "e-NMU(v)",
  "Ctrl(i)", "Ctrl(ii)", "Ctrl(iii)", "Ctrl(iv)", "Ctrl(v)"
)

# ── per (element, slot): mean raw activity in each orientation, side by side ─
ori_scatter_data <- ori_long_scatter %>%
  group_by(element_id, element_label, slot, orientation) %>%
  summarise(mean_raw = mean(avg_Rep), n = n(), .groups = "drop") %>%
  pivot_wider(names_from = orientation, values_from = c(mean_raw, n)) %>%
  mutate(
    element_label = factor(as.character(element_label), levels = fc_levels_enmu_top)
  ) %>%
  filter(!is.na(element_label), !is.na(mean_raw_forward), !is.na(mean_raw_reverse))

cat("── n per (element, slot) point ──\n")
print(ori_scatter_data %>% select(element_label, slot, n_forward, n_reverse), n = Inf)

# ── slot color scale (distinct from the element/family palettes above) ──────
SLOT_COLORS <- c("1" = "#C03830", "2" = "#E7872B", "3" = "#E9C61D",
                 "4" = "#356920", "5" = "#2D5F8A")

# ── shared axis limits — tight to actual data range, no manual padding ─────
axis_lo <- min(ori_scatter_data$mean_raw_forward + 0.05,
              ori_scatter_data$mean_raw_reverse + 0.05)
axis_hi <- max(ori_scatter_data$mean_raw_forward + 0.05,
              ori_scatter_data$mean_raw_reverse + 0.05)

cat(sprintf("Data range (shared axis): %.4f – %.4f\n", axis_lo, axis_hi))

# ── only keep candidate breaks (10^-1, 10^0, 10^1) that fall within range ──
candidate_breaks <- 10^(-1:1)
axis_breaks <- candidate_breaks[candidate_breaks >= axis_lo & candidate_breaks <= axis_hi]

cat("Breaks shown:", paste(axis_breaks, collapse = ", "), "\n")

# ── All 10 elements (e-NMU top row, Ctrl bottom row), by slot ───────────────
p_orientation_scatter_all <- ggplot(ori_scatter_data,
       aes(x = mean_raw_forward + 0.05, y = mean_raw_reverse + 0.05,
           color = factor(slot))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  geom_point(size = 2, alpha = 0.85) +
  scale_color_manual(values = SLOT_COLORS, name = "Slot", labels = paste0("S", 1:5)) +
  scale_x_log10(breaks = axis_breaks, labels = scales::label_log(),
                limits = c(axis_lo, axis_hi)) +
  scale_y_log10(breaks = axis_breaks, labels = scales::label_log(),
                limits = c(axis_lo, axis_hi)) +
  coord_fixed(ratio = 1) +
  annotation_logticks(sides = "bl", size = 0.3) +
  facet_wrap(~ element_label, nrow = 2) +
  labs(
    x = "Mean activity - forward orientation",
    y = "Mean activity - reverse orientation"
  ) +
  theme(
    strip.text        = element_text(size = 8),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank()
  )

print(p_orientation_scatter_all)
save_fig("figS2d_orientation_scatter_fwd_rev", p_orientation_scatter_all,
         panel_width = PANEL_W * 0.64, panel_height = PANEL_H * 0.5)
```

# Supplementary Figure 3
## Fig S3a
```{r}
id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

df_insu_count <- df %>%
  filter(S5 == "e-NMU(iii)F") %>%
  rowwise() %>%
  mutate(
    n_insu = sum(c_across(all_of(id_cols_s1s4)) == "synthetic_insulator")
  ) %>%
  ungroup() %>%
  mutate(
    n_insu_label = factor(n_insu, levels = 0:4)
  )

df_insu_count_summary <- df_insu_count %>%
  mutate(log_act = log10(avg_Rep + 0.05)) %>%
  group_by(n_insu_label) %>%
  summarise(
    mean_log = mean(log_act),
    se_log   = sd(log_act) / sqrt(n()),
    ci95_log = se_log * 1.96,
    n        = n(),
    .groups  = "drop"
  )

p_insu_violin <- ggplot(df_insu_count,
                         aes(x = n_insu_label, y = avg_Rep + 0.05)) +
  geom_violin(trim = FALSE, alpha = 0.7,
              fill = unname(COLOR_INSULATOR), color = "grey30") +
  geom_boxplot(width = 0.1, fill = unname(COLOR_INSULATOR), color = "grey30",
               outlier.shape = NA) +
  geom_text(data = df_insu_count_summary,
            aes(x = n_insu_label, y = Inf, label = paste0("n = ", scales::comma(n))),
            inherit.aes = FALSE,
            vjust = -0.4, size = 3, family = FONT) +
  scale_y_log10(labels = scales::label_log()) +
  annotation_logticks(sides = "l", size = 0.3) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Insulator copy number (S1–S4)",
    y = "Activity score"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin      = margin(t = 20, r = 10, b = 5, l = 5)
  )

print(p_insu_violin)
save_fig("figS3a_insu_count_violin", p_insu_violin, panel_width = PANEL_W*1.6, panel_height = PANEL_H)

mean_0 <- df_insu_count_summary$mean_log[df_insu_count_summary$n_insu_label == "0"]
mean_1 <- df_insu_count_summary$mean_log[df_insu_count_summary$n_insu_label == "1"]
n_0    <- df_insu_count_summary$n[df_insu_count_summary$n_insu_label == "0"]
n_1    <- df_insu_count_summary$n[df_insu_count_summary$n_insu_label == "1"]

fold_diff <- 10^(mean_0 - mean_1)

cat(sprintf("Constructs with 0 insulators (n = %d) show a %.2f-fold difference in activity relative to constructs with 1 insulator (n = %d).\n",
            n_0, fold_diff, n_1))

# ── 0 insulators vs. any (≥1) insulator ──────────────────────────────────────
mean_log_any_insu <- df_insu_count %>%
  filter(n_insu >= 1) %>%
  summarise(mean_log = mean(log10(avg_Rep + 0.05))) %>%
  pull(mean_log)

fold_diff_any <- 10^(mean_0 - mean_log_any_insu)

cat(sprintf("Constructs with 0 insulators (n = %d) show a %.2f-fold difference in activity relative to constructs with any insulator present (1–4, pooled, n = %d).\n",
            n_0, fold_diff_any, sum(df_insu_count$n_insu >= 1)))
```

## Fig S3b and S3c
```{r}
id_cols_s1s4 <- paste0("insert_", 1:4, "_id")

# ── shared filtered subset ───────────────────────────────────────────────────
df_base_de <- df %>%
  filter(S5 == "e-NMU(iii)F") %>%
  rowwise() %>%
  mutate(
    n_insu       = sum(c_across(all_of(id_cols_s1s4)) == "synthetic_insulator"),
    n_enmu_iiiiv = sum(str_detect(c_across(all_of(id_cols_s1s4)), "eNMU_region_(iii|iv)")),
    n_enmu_any   = sum(str_detect(c_across(all_of(id_cols_s1s4)), "eNMU_region"))
  ) %>%
  ungroup() %>%
  filter(n_insu == 1, n_enmu_iiiiv == 1, n_enmu_any == 1) %>%
  rowwise() %>%
  mutate(
    insu_slot = which(c_across(all_of(id_cols_s1s4)) == "synthetic_insulator"),
    enmu_slot = which(str_detect(c_across(all_of(id_cols_s1s4)), "eNMU_region_(iii|iv)"))
  ) %>%
  ungroup() %>%
  filter(insu_slot != enmu_slot) %>%
  mutate(
    enmu_side = if_else(enmu_slot < insu_slot, "eNMU left of insulator",
                                               "eNMU right of insulator"),
    enmu_side = factor(enmu_side, levels = c("eNMU right of insulator",
                                             "eNMU left of insulator")),
    log_act   = log10(avg_Rep + 0.05)
  )

cat(nrow(df_base_de), "constructs after filtering.\n")

# ── Fig 2d — grouped: left vs. right of insulator ────────────────────────────
grouped_summary_2d <- df_base_de %>%
  group_by(enmu_side) %>%
  summarise(mean_log = mean(log_act), n = n(), .groups = "drop")

label_map_2d    <- setNames(paste0(grouped_summary_2d$enmu_side, " (n = ", grouped_summary_2d$n, ")"),
                            grouped_summary_2d$enmu_side)
label_levels_2d <- label_map_2d[levels(df_base_de$enmu_side)]

df_2d <- df_base_de %>%
  mutate(label = factor(label_map_2d[as.character(enmu_side)], levels = label_levels_2d))

grouped_summary_2d <- grouped_summary_2d %>%
  mutate(
    label   = factor(label_map_2d[as.character(enmu_side)], levels = label_levels_2d),
    x_num   = as.numeric(label)
  )

p_2d_beeswarm <- ggplot(df_2d, aes(x = label, y = 10^log_act)) +
  geom_violin(trim = FALSE, fill = unname(COLOR_OTHER), alpha = 0.7,
             color = "grey30", linewidth = 0.4, scale = "width") +
  geom_quasirandom(size = 1.1, alpha = 0.5, width = 0.28,
                   color = "#3B3B3B", stroke = 0) +
  geom_segment(
    data = grouped_summary_2d,
    aes(x = x_num - 0.4, xend = x_num + 0.4, y = 10^mean_log, yend = 10^mean_log),
    inherit.aes = FALSE, color = "grey30", linewidth = 0.9
  ) +
  scale_y_log10(breaks = c(1, 10, 100), labels = scales::label_log()) +
  annotation_logticks(sides = "b", size = 0.3) +
  coord_flip() +
  labs(x = NULL, y = "Activity score") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y       = element_text(family = FONT, size = 9)
  )

print(p_2d_beeswarm)
save_fig("figS3b_insu_lr_beeswarm_violin_all", p_2d_beeswarm, panel_width = PANEL_W * 0.8, panel_height = PANEL_H)

cat("\n── Fig 2d group n / mean ──\n")
print(grouped_summary_2d %>% select(enmu_side, n, mean_log))

# ── Fig 2e — individual (insu_slot × enmu_slot) combinations ────────────────
combo_summary_2e <- df_base_de %>%
  group_by(enmu_side, insu_slot, enmu_slot) %>%
  summarise(mean_log = mean(log_act), n = n(), .groups = "drop") %>%
  mutate(
    label = pmap_chr(
      list(insu_slot, enmu_slot, n),
      function(ins, enm, nn) {
        slot_contents <- sapply(1:4, function(s) {
          if (s == ins) "Insu"
          else if (s == enm) "eNMU"
          else "·"
        })
        sprintf("[%s]      (n = %d)", paste(slot_contents, collapse = " "), nn)
      }
    ),
    sort_key = if_else(enmu_slot < insu_slot, 0, 1) * 100 + insu_slot * 10 + enmu_slot,
    label    = fct_reorder(label, -sort_key)
  )

df_2e <- df_base_de %>%
  left_join(combo_summary_2e %>% select(insu_slot, enmu_slot, label),
            by = c("insu_slot", "enmu_slot"))

combo_summary_2e <- combo_summary_2e %>%
  mutate(x_num = as.numeric(label))

p_2e_beeswarm <- ggplot(df_2e, aes(x = label, y = 10^log_act)) +
  geom_violin(trim = FALSE, fill = unname(COLOR_OTHER), alpha = 0.7,
             color = "grey30", linewidth = 0.35, scale = "width") +
  geom_quasirandom(size = 0.9, alpha = 0.5, width = 0.28,
                   color = "#3B3B3B", stroke = 0) +
  geom_segment(
    data = combo_summary_2e,
    aes(x = x_num - 0.4, xend = x_num + 0.4, y = 10^mean_log, yend = 10^mean_log),
    inherit.aes = FALSE, color = "grey30", linewidth = 0.8
  ) +
  scale_y_log10(breaks = c(1, 10, 100), labels = scales::label_log()) +
  annotation_logticks(sides = "b", size = 0.3) +
  coord_flip() +
  labs(x = NULL, y = "Activity score") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(family = FONT, size = 9)
  )

print(p_2e_beeswarm)
save_fig("figS3c_insu_lr_beeswarm_violin", p_2e_beeswarm, panel_width = PANEL_W * 2.2, panel_height = PANEL_H * 1.3)

cat("\n── Fig 2e combo n / mean ──\n")
print(combo_summary_2e %>% select(enmu_side, insu_slot, enmu_slot, n, mean_log))
```

## Fig S3d
```{r}
insu_scatter_data <- ori_long_scatter %>%
  filter(element_label == "Insu(i)") %>%
  group_by(element_id, element_label, slot, orientation) %>%
  summarise(mean_raw = mean(avg_Rep), n = n(), .groups = "drop") %>%
  pivot_wider(names_from = orientation, values_from = c(mean_raw, n)) %>%
  filter(!is.na(mean_raw_forward), !is.na(mean_raw_reverse))

cat("── n per (Insu(i), slot) point ──\n")
print(insu_scatter_data %>% select(slot, n_forward, n_reverse), n = Inf)

p_orientation_scatter_insu <- ggplot(insu_scatter_data,
       aes(x = mean_raw_forward + 0.05, y = mean_raw_reverse + 0.05,
           color = factor(slot))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  geom_point(size = 2, alpha = 0.85) +
  scale_color_manual(values = SLOT_COLORS, name = "Slot", labels = paste0("S", 1:5)) +
  scale_x_log10(breaks = axis_breaks, labels = scales::label_log(),
                limits = c(axis_lo, axis_hi)) +
  scale_y_log10(breaks = axis_breaks, labels = scales::label_log(),
                limits = c(axis_lo, axis_hi)) +
  coord_fixed(ratio = 1) +
  annotation_logticks(sides = "bl", size = 0.3) +
  facet_wrap(~ element_label) +
  labs(
    x = "Mean activity - forward orientation",
    y = "Mean activity - reverse orientation"
  ) +
  theme(
    strip.text        = element_text(size = 8),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank()
  )

print(p_orientation_scatter_insu)

save_fig("figS3d_orientation_scatter_insu",
         p_orientation_scatter_insu,
         panel_width = PANEL_W, panel_height = PANEL_H)
```
