---
title: "LAMPRA Paper Figures — Fig1 & FigS1"
author: "cbiar"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    fig_width: 10
    fig_height: 10
  pdf_document:
    toc: true
    toc_depth: '3'
editor_options:
  chunk_output_type: console
---

<!--
OVERVIEW
========
  Fig 1c - Read counts per insert combination (histogram)
  Fig 1d - Replicate-vs-replicate scatter
  Fig 1e - LAMPRA activity violin by promoter group 
  
  Fig S1b - Total read counts per insert, forward vs reverse (bar)
  Fig S1c - Element distribution by slot (stacked bar)
  Fig S1d - LAMPRA activity violin by S5 identity 

HOW TO RUN
  1. Open this file from the project's RStudio .Rproj (or set the working
     directory to the project root) so that here::here() resolves correctly.
  2. Make sure the input files in the "Input Paths" chunk exist under
     the project root.
  3. Knit, or run chunks top-to-bottom. Chunk order matters.
  4. Figures are written to ./figures-fig1-figS1/ as PDFs.

DEPENDENCIES
  CRAN: tidyverse, scales, RColorBrewer, cowplot, sysfonts, showtext, here.
  Fonts: the setup chunk loads Arial from macOS paths — see the note there if
  you're on Linux/Windows.
-->

```{r setup, include=FALSE}
# ── Output directory ──
FIG_DIR <- file.path(here::here(), "figures-fig1-figS1")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 10,
  dpi = 150,
  dev = c("pdf", "png"),
  fig.path = paste0(FIG_DIR, "/")
)
```

# Setup

## Libraries

```{r libraries}
library(tidyverse)
library(scales)       
library(RColorBrewer) 
library(cowplot)      
library(sysfonts)     
library(showtext)     
library(egg)
library(gtable)
library(grid)

# ── Load Arial for PDF/PNG output ──
# NOTE: these paths are macOS-specific. On Linux/Windows, point font_add() at a
# local Arial .ttf (or swap in any installed font and set FONT below to match).
# If you don't care about the exact typeface, you can delete this block and the
# showtext_auto() call — plots will fall back to the default device font.
font_add("Arial", regular = "/Library/Fonts/Arial.ttf",
         bold = "/Library/Fonts/Arial Bold.ttf",
         italic = "/Library/Fonts/Arial Italic.ttf",
         bolditalic = "/Library/Fonts/Arial Bold Italic.ttf")

showtext_auto()

# Shorthand for text geoms
FONT <- "Arial"

# ── Project theme ──
pub_theme <- function() {
  theme_cowplot(font_size = 8) %+replace%
    theme(
      strip.background = element_rect(color = 'black',
                                      fill = 'grey90',
                                      linetype = "solid",
                                      linewidth = 0.25),
      text             = element_text(size = 8, family = FONT),
      axis.text.x      = element_text(size = 8),
      axis.text.y      = element_text(size = 8, hjust = 1, margin = margin(r = 2)),
      axis.title.x      = element_text(size = 9, margin = margin(t = -1)),
      axis.title.y      = element_text(size = 9, angle = 90, margin = margin(r = -1)),
      axis.line         = element_line(linewidth = 0.25),
      strip.text        = element_text(family = FONT, size = 8, face = 'bold',
                                       margin = margin(b = 2.5, t = 2.5)),
      legend.text       = element_text(size = 8),
      legend.title      = element_text(size = 8),
      legend.position   = 'right',
      legend.title.align = 0.5,
      plot.tag          = element_text(size = 14, face = 'bold'),
      plot.title        = element_text(hjust = 0.5, face = 'plain', size = 12,
                                       margin = margin(b = 10))
    )
}
theme_set(pub_theme())

# ── Panel dimension constants ─────────────────────────────────────────────────
PANEL_W      <- unit(2.3, "in")
PANEL_H      <- unit(2.3, "in")

# ── Annotation text size (matches theme's 10pt in ggplot2 size units) ─────────
ANNOT_SIZE <- 8 / .pt   # ≈ 3.53 — use this in annotate() and stat_summary()

# ── Log10 axis helper ─────────────────────────────────────────────────────────
# Applies consistent breaks (10^-2 through 10^3), 10^x labels, and logticks
# to whichever axis you specify. Use like:
#   p + log_scale("x") + log_scale("y")
# The `limits` arg is optional — pass c(lo, hi) in raw (untransformed) units.
LOG_BREAKS <- 10^(-2:3)   # 0.01, 0.1, 1, 10, 100, 1000

log_scale <- function(axis = c("x", "y"), limits = NULL, expand = waiver(), breaks = LOG_BREAKS) {
  axis <- match.arg(axis)
  args <- list(
    trans   = "log10",
    breaks  = breaks,
    labels  = scales::label_log(base = 10),
    limits  = limits,
    expand  = expand
  )
  if (axis == "x") do.call(scale_x_continuous, args)
  else             do.call(scale_y_continuous, args)
}

# ── save_fig() helper ─────────────────────────────────────────────────────────
# Wraps ggsave, always writing to FIG_DIR.  Call like:
#   save_fig("Fig1_scatter", p, width = 7, height = 5)
save_fig <- function(name, plot = last_plot(),
                     panel_width  = PANEL_W,
                     panel_height = PANEL_H,
                     dpi = 300, ext = "pdf", ...) {
  fpath <- file.path(FIG_DIR, paste0(name, ".", ext))

  sized   <- egg::set_panel_size(plot, width = panel_width, height = panel_height)
  total_w <- grid::convertWidth( gtable::gtable_width(sized),  "in", valueOnly = TRUE)
  total_h <- grid::convertHeight(gtable::gtable_height(sized), "in", valueOnly = TRUE)

  if (ext == "pdf") {
    quartz(type = "pdf", file = fpath,
           width = total_w, height = total_h, family = "Arial")
    grid::grid.draw(sized)
    dev.off()
  } else if (ext == "png") {
    quartz(type = "png", file = fpath,
           width = total_w * dpi, height = total_h * dpi, family = "Arial")
    grid::grid.draw(sized)
    dev.off()
  }

  if (file.exists(fpath) && file.size(fpath) > 0) {
    message(sprintf("Saved: %s  [%.2f × %.2f in]", fpath, total_w, total_h))
  } else {
    warning(sprintf("FAILED to write: %s", fpath))
  }
  invisible(fpath)
}
```

## Input Paths

This subset reads these data files (all expected under the project root,
resolved with `here::here()` — run from the project's `.Rproj` or set the
working directory there).

| Variable      | File                                                        
|---------------|-------------------------------------------------------------
| `RAW_DATA`    | `data/raw/v2_longMPRA_scores_with_orientation_20260127.txt` 
| `PACBIO_DATA` | `data/association/seq008_longMPRA_v1_association_df.txt.gz`                 
| `PACBIO_MIN`  | `data/association/seq008_longMPRA_v1_association_minimal.txt.gz`            

```{r input-paths}
# ── Project root ──
PROJECT <- here::here()

# ── Raw LAMPRA scores ──
RAW_DATA    <- file.path(PROJECT, "data/raw/v2_longMPRA_scores_with_orientation_20260127.txt")

# ── PacBio barcode-association data ──
PACBIO_DATA <- file.path(PROJECT, "data/association/seq008_longMPRA_v1_association_df.txt.gz")
PACBIO_MIN  <- file.path(PROJECT, "data/association/seq008_longMPRA_v1_association_minimal.txt.gz")
```

## Color Palette

```{r colors}
# RColorBrewer Set2 — 8 colors
SET2 <- brewer.pal(8, "Set2")
names(SET2) <- c("teal", "orange", "lavender", "pink",
                 "yellow_green", "yellow", "tan", "gray")
# 1  #66C2A5  teal
# 2  #FC8D62  orange
# 3  #8DA0CB  lavender
# 4  #E78AC3  pink
# 5  #A6D854  yellow-green
# 6  #FFD92F  yellow
# 7  #E5C494  tan
# 8  #B3B3B3  gray

# ── Semantic color assignments ──
COLOR_ORIENTATION <- paste0(SET2["pink"])   # forward/reverse bars
COLOR_INSULATOR   <- SET2["orange"]         # synthetic insulator


# ── Consolidated palettes (v9) ────────────────────────────────────────────────
# Every static color palette used downstream lives here. Data-derived color
# maps (e.g. part_color_map built from slot_shap_data) stay where they're
# computed.

# Sequential teal ramp for the five e-NMU regions, anchored so the Set2 teal
# base lands exactly at step 3 (e-NMU(iii)).
# Knobs:
#   ENMU_LIGHT / ENMU_DARK widen or narrow the ramp endpoints.
#   ENMU_OVERRIDES nudges any individual step — set to a hex to override.
ENMU_LIGHT <- "#DCEFE7"   # e-NMU(i) target before overrides
ENMU_BASE  <- "#66C2A5"   # Set2 teal — fixed at step 3 (e-NMU(iii))
ENMU_DARK  <- "#0A4A33"   # e-NMU(v) target before overrides
enmu_palette <- colorRampPalette(c(ENMU_LIGHT, ENMU_BASE, ENMU_DARK))(5)
# Per-step overrides. Order: i, ii, iii, iv, v. NA = keep the ramp value.
ENMU_OVERRIDES <- c(NA, NA, NA, NA, NA)

enmu_palette   <- ifelse(is.na(ENMU_OVERRIDES), enmu_palette, ENMU_OVERRIDES)


COLOR_IIIF  <- SET2["teal"]    # #66C2A5
COLOR_OTHER <- "gray"
COLOR_IVRF <- enmu_palette[4]   


# Sequential blue ramp for the five Ctrl (dinuc-shuffled) regions, anchored
# so the Set2 lavender base lands exactly at step 3 (Ctrl(iii)). Dark end is
# pulled toward navy to keep the ramp reading as blue, not purple.
# Knobs:
#   CTRL_LIGHT / CTRL_DARK widen or narrow the ramp endpoints.
#   CTRL_OVERRIDES nudges any individual step — set to a hex to override.
CTRL_LIGHT <- "#E5EAF4"   # Ctrl(i) target before overrides
CTRL_BASE  <- "#8DA0CB"   # Set2 lavender — fixed at step 3 (Ctrl(iii))
CTRL_DARK  <- "#1F2C5C"   # Ctrl(v) target before overrides
ctrl_palette <- colorRampPalette(c(CTRL_LIGHT, CTRL_BASE, CTRL_DARK))(5)
# Per-step overrides. Order: i, ii, iii, iv, v. NA = keep the ramp value.
CTRL_OVERRIDES <- c(NA, NA, NA, NA, NA)

ctrl_palette   <- ifelse(is.na(CTRL_OVERRIDES), ctrl_palette, CTRL_OVERRIDES)

# 11-entry map keyed by raw insert_type strings — used by the Fig S1c
# stacked-bar plot (enmu_palette / ctrl_palette feed it).
color_map_detail <- c(
  synthetic_insulator       = unname(COLOR_INSULATOR),
  eNMU_region_i             = enmu_palette[1],
  eNMU_region_ii            = enmu_palette[2],
  eNMU_region_iii           = enmu_palette[3],
  eNMU_region_iv            = enmu_palette[4],
  eNMU_region_v             = enmu_palette[5],
  dinucleotide_shuffled_i   = ctrl_palette[1],
  dinucleotide_shuffled_ii  = ctrl_palette[2],
  dinucleotide_shuffled_iii = ctrl_palette[3],
  dinucleotide_shuffled_iv  = ctrl_palette[4],
  dinucleotide_shuffled_v   = ctrl_palette[5]
)

COLOR_IIIR        <- enmu_palette[3]  
COLOR_ENMU_OTHER  <- enmu_palette[2]  
COLOR_CTRL_ALL    <- unname(CTRL_BASE)
COLOR_INSU_ALL    <- unname(COLOR_INSULATOR)

# Explicit display order for stacked_detail: insulator → e-NMU i–v → Ctrl i–v.
type_order_types  <- c(
  "synthetic_insulator",
  "eNMU_region_i", "eNMU_region_ii", "eNMU_region_iii",
  "eNMU_region_iv", "eNMU_region_v",
  "dinucleotide_shuffled_i",   "dinucleotide_shuffled_ii",
  "dinucleotide_shuffled_iii", "dinucleotide_shuffled_iv",
  "dinucleotide_shuffled_v"
)
type_order_labels <- c(
  "Insu(i)",
  "e-NMU(i)", "e-NMU(ii)", "e-NMU(iii)", "e-NMU(iv)", "e-NMU(v)",
  "Ctrl(i)",  "Ctrl(ii)",  "Ctrl(iii)",  "Ctrl(iv)",  "Ctrl(v)"
)

# Orientation map (forward / reverse) for BC-level count bars.
custom_colors <- c("forward"  = COLOR_ORIENTATION,
                   "reverse"  = paste0(COLOR_ORIENTATION, "99"))

# Label shortener: maps raw feature/insert names to display labels.
# Used by Fig S1b to relabel the x-axis insert names.
short_label <- function(name) {
  name %>%
    # Extract and reformat slot number to append at end
    str_replace_all("insert_([1-5])_id_eNMU_region_([ivx]+)", "e-NMU(\\2)-S\\1") %>%
    str_replace_all("insert_([1-5])_id_dinucleotide_shuffled_([ivx]+)", "Ctrl(\\2)-S\\1") %>%
    str_replace_all("insert_([1-5])_id_synthetic_insulator", "Insu(i)-S\\1") %>%
    str_replace_all("insert_([1-5])_id_eNMU_region", "e-NMU-S\\1") %>%   # fallback no numeral
    str_replace_all("insert_([1-5])_id_dinucleotide_shuffled", "Ctrl-S\\1") %>%  # fallback
    str_replace_all("insert_([1-5])_ori_reverse", "ori-S\\1") %>%
    str_replace_all("insert_([1-5])_", "S\\1 ")   # catch-all for anything remaining
}
```

## PacBio association data prep

This chunk loads `association_df`, splits `insert_combo` into per-slot columns,
and computes the BC-level `table2_bc_counts` and the per-combo `combo_counts`
used downstream by count histogram and forward/reverse barplot.

```{r pacbio-association-prep}
######################################
# Read in data
######################################
# Full association df (insert_combo, BC, count)
association_df = read.table(PACBIO_DATA, header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

# Minimal association df (BC, insert_combo only - no counts)
association_minimal <- read.table(PACBIO_MIN,
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE
)

# Number of unique insert combos
n_distinct(association_df$insert_combo)
n_distinct(association_minimal$insert_combo)


######################################
# Split insert_combo into individual columns
######################################

insert_cols <- c("insert_1", "insert_2", "insert_3", "insert_4", "insert_5")

association_df <- association_df %>%
  separate(insert_combo, into = insert_cols, sep = ";", remove = FALSE)

######################################
# Table 1: Number of unique combos containing each insert (construct level)
######################################

table1_construct_combos <- lapply(insert_cols, function(col) {
  association_df %>%
    group_by(insert = .data[[col]]) %>%
    summarise(n_unique_combos = n_distinct(insert_combo), .groups = "drop")
}) %>%
  bind_rows() %>%
  group_by(insert) %>%
  summarise(n_unique_combos = sum(n_unique_combos), .groups = "drop") %>%
  mutate(
    base_name = gsub("_forward|_reverse", "", insert),
    orientation = case_when(
      grepl("_forward$", insert) ~ "forward",
      grepl("_reverse$", insert) ~ "reverse",
      TRUE ~ "unmapped"
    )
  ) %>%
  arrange(desc(n_unique_combos))

print("Table 1: Unique insert combos per insert (construct level)")
print(table1_construct_combos)

######################################
# Table 2: Sum of counts per insert (BC level)
######################################

table2_bc_counts <- lapply(insert_cols, function(col) {
  association_df %>%
    group_by(insert = .data[[col]]) %>%
    summarise(total_counts = sum(count), .groups = "drop")
}) %>%
  bind_rows() %>%
  group_by(insert) %>%
  summarise(total_counts = sum(total_counts), .groups = "drop") %>%
  mutate(
    base_name = gsub("_forward|_reverse", "", insert),
    orientation = case_when(
      grepl("_forward$", insert) ~ "forward",
      grepl("_reverse$", insert) ~ "reverse",
      TRUE ~ "unmapped"
    )
  ) %>%
  arrange(desc(total_counts))

print("Table 2: Total read counts per insert (BC level)")
print(table2_bc_counts)

######################################
# Table 3: Overall forward vs reverse at construct level
######################################

table3_orientation_construct <- table1_construct_combos %>%
  group_by(orientation) %>%
  summarise(n_unique_combos = sum(n_unique_combos), .groups = "drop") %>%
  mutate(pct = round(n_unique_combos / sum(n_unique_combos) * 100, 1))

print("Table 3: Overall forward vs reverse at construct level")
print(table3_orientation_construct)

######################################
# Table 4: Overall forward vs reverse at BC level
######################################

table4_orientation_bc <- table2_bc_counts %>%
  group_by(orientation) %>%
  summarise(total_counts = sum(total_counts), .groups = "drop") %>%
  mutate(pct = round(total_counts / sum(total_counts) * 100, 1))

print("Table 4: Overall forward vs reverse at BC level")
print(table4_orientation_bc)

######################################
# Fold range of inserts (BC level)
######################################
# Sum forward + reverse per base insert
insert_totals_bc <- table2_bc_counts %>%
  group_by(base_name) %>%
  summarise(total_counts = sum(total_counts), .groups = "drop") %>%
  arrange(desc(total_counts))

print("Insert totals at BC level:")
print(insert_totals_bc)

# Fold range: highest / lowest
max_insert <- insert_totals_bc %>% slice_max(total_counts, n = 1)
min_insert <- insert_totals_bc %>% slice_min(total_counts, n = 1)

fold_range_bc <- max_insert$total_counts / min_insert$total_counts

cat("\nHighest:", max_insert$base_name, "-", max_insert$total_counts, "counts\n")
cat("Lowest: ", min_insert$base_name, "-", min_insert$total_counts, "counts\n")
cat("Fold range:", round(fold_range_bc, 2), "\n")

######################################
# combo_counts: total counts per unique insert combo (used by Fig 1c histogram)
######################################
combo_counts <- association_df %>%
  group_by(insert_combo) %>%
  summarise(total_counts = sum(count), .groups = "drop")

######################################
# Calculate how much data within 4 and 10 fold range of median and mean
######################################

median_count <- median(combo_counts$total_counts)
mean_count <- mean(combo_counts$total_counts)
n_total <- nrow(combo_counts)

# 4-fold range (2x either side)
cat("--- 4-fold range ---\n")
cat("Around median [", median_count/2, ",", median_count*2, "]\n")
cat(sum(combo_counts$total_counts >= median_count/2 & combo_counts$total_counts <= median_count*2),
    "/", n_total, "(", round(mean(combo_counts$total_counts >= median_count/2 & combo_counts$total_counts <= median_count*2)*100, 1), "%)\n")

cat("Around mean [", mean_count/2, ",", mean_count*2, "]\n")
cat(sum(combo_counts$total_counts >= mean_count/2 & combo_counts$total_counts <= mean_count*2),
    "/", n_total, "(", round(mean(combo_counts$total_counts >= mean_count/2 & combo_counts$total_counts <= mean_count*2)*100, 1), "%)\n")

# 10-fold range (sqrt(10) either side on log scale)
cat("\n--- 10-fold range ---\n")
cat("Around median [", median_count/sqrt(10), ",", median_count*sqrt(10), "]\n")
cat(sum(combo_counts$total_counts >= median_count/sqrt(10) & combo_counts$total_counts <= median_count*sqrt(10)),
    "/", n_total, "(", round(mean(combo_counts$total_counts >= median_count/sqrt(10) & combo_counts$total_counts <= median_count*sqrt(10))*100, 1), "%)\n")

cat("Around mean [", mean_count/sqrt(10), ",", mean_count*sqrt(10), "]\n")
cat(sum(combo_counts$total_counts >= mean_count/sqrt(10) & combo_counts$total_counts <= mean_count*sqrt(10)),
    "/", n_total, "(", round(mean(combo_counts$total_counts >= mean_count/sqrt(10) & combo_counts$total_counts <= mean_count*sqrt(10))*100, 1), "%)\n")
```

## Raw LAMPRA score data 
```{r load-raw-data}
df_raw <- read.table(RAW_DATA, header = TRUE,  sep = "\t",
                     stringsAsFactors = FALSE)

# add pseudocount, log transform
df <- df_raw %>%
  mutate(
    S5 = case_when(
      insert_5 == "eNMU_region_iii_forward" ~ "e-NMU(iii)F",
      insert_5 == "eNMU_region_iv_reverse"  ~ "e-NMU(iv)R",
      TRUE                                  ~ "Other"
    ),
    RepA_log05 = log10(RepA + 0.05),
    RepB_log05 = log10(RepB + 0.05),
    RepC_log05 = log10(RepC + 0.05),
    avg_Rep_log05 = log10(avg_Rep + 0.05)
  )

# Mean with log score ->
df %>%
  group_by(S5) %>%
  summarise(mean_activity = mean(avg_Rep_log05, na.rm = TRUE),
            se = sd(avg_Rep_log05, na.rm = TRUE) / sqrt(n()),
            n = n())

# REPORTED ACTIVITY SCORE MEANS AND RANGES
df %>%
  group_by(S5) %>%
  summarise(mean_activity = mean(avg_Rep, na.rm = TRUE),
            sd = sd(avg_Rep, na.rm = TRUE),
            se = sd(avg_Rep, na.rm = TRUE) / sqrt(n()),
            n = n(),
            min = min(avg_Rep, na.rm = TRUE),
            max = max(avg_Rep, na.rm = TRUE))
```

# Figure 1
## Fig 1c
```{r fig1c-combo-uniformity}
log_breaks <- 10^seq(0, 3.1, length.out = 51)

plt_combo_uniformity <- ggplot(combo_counts, aes(x = total_counts)) +
  stat_bin(geom = "step", breaks = log_breaks, color = "black", linewidth = 0.5) +
  log_scale("x", breaks = 10^(0:4), limits = c(1, 10^4)) +
  annotation_logticks(sides = "b", size = 0.3) +
  scale_y_continuous(limits = c(0, 3000), breaks = c(0, 1000, 2000, 3000)) +
  labs(
    x = "Reads per sCRL",
    y = "Number of sCRLs"
  )
print(plt_combo_uniformity)

# Export for publication
save_fig("fig1c_insert_combo_uniformity", plt_combo_uniformity, panel_height = PANEL_H * 0.9, panel_width = PANEL_W * 0.9 )
```

## Fig 1d
```{r}
df <- df %>%
  mutate(
    S5 = case_when(
      insert_5 == "eNMU_region_iii_forward" ~ "e-NMU(iii)F",
      insert_5 == "eNMU_region_iii_reverse" ~ "e-NMU(iii)R",
      insert_5 == "eNMU_region_iv_reverse"  ~ "e-NMU(iv)R",
      TRUE                                  ~ "All other"
    ),
    S5 = factor(S5, levels = c("e-NMU(iii)F", "e-NMU(iii)R", "e-NMU(iv)R", "All other"))
  )

r2_BC <- cor(
  df %>% filter(!is.na(RepB_log05), !is.na(RepC_log05)) %>% pull(RepB_log05),
  df %>% filter(!is.na(RepB_log05), !is.na(RepC_log05)) %>% pull(RepC_log05),
  method = "pearson"
)

df_all <- df %>% filter(!is.na(RepB_log05), !is.na(RepC_log05))

set.seed(42)
df_shuffled <- df_all %>% slice_sample(prop = 1)

r_label <- sprintf('italic(r) ~ "= %.2f"', r2_BC)

base_scales <- list(
  log_scale("x"),
  log_scale("y"),
  annotation_logticks(sides = "bl", size = 0.3),
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black"),
  labs(x = "Replicate 1 activity score", y = "Replicate 2 activity score"),
  annotate("text", x = 3.5, y = Inf, hjust = 0.5, vjust = 1.5,
           label = r_label, parse = TRUE, size = ANNOT_SIZE),
  theme(
    legend.position      = "bottom",
    legend.justification = "center",
    panel.grid.major     = element_blank(),
    panel.grid.minor     = element_blank()
  )
)

plt_rep_scatter_iiiR <- ggplot(df_shuffled,
       aes(x = RepB + 0.05, y = RepC + 0.05, color = S5,
           shape = S5, size = S5, alpha = S5, stroke = S5)) +
  geom_point() +
  scale_color_manual(
    values = c(
      "All other"        = unname(COLOR_OTHER),
      "e-NMU(iv)R"  = unname(COLOR_IVRF),
      "e-NMU(iii)F" = unname(COLOR_IIIF),
      "e-NMU(iii)R" = unname(COLOR_IIIR)
    ),
    name = "S5 identity"
  ) +
  scale_shape_manual(
    values = c("All other" = 16, "e-NMU(iv)R" = 16,
               "e-NMU(iii)F" = 16, "e-NMU(iii)R" = 4),
    name = "S5 identity"
  ) +
  scale_size_manual(
    values = c("All other" = 0.5, "e-NMU(iv)R" = 0.5,
               "e-NMU(iii)F" = 0.5, "e-NMU(iii)R" = 0.5),
    name = "S5 identity"
  ) +
  scale_alpha_manual(
    values = c("All other" = 0.7, "e-NMU(iv)R" = 0.7,
               "e-NMU(iii)F" = 0.7, "e-NMU(iii)R" = 0.7),
    name = "S5 identity"
  ) +
  scale_discrete_manual("stroke",
    values = c("All other" = 0.3, "e-NMU(iv)R" = 0.3,
               "e-NMU(iii)F" = 0.3, "e-NMU(iii)R" = 0.3),
    name = "S5 identity"
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 1.5), nrow = 2),
    shape = guide_legend(nrow = 2),
    alpha = guide_legend(nrow = 2),
    size  = guide_legend(nrow = 2)
  ) +
  base_scales

print(plt_rep_scatter_iiiR)
save_fig("fig1d_replicate_scatter", plt_rep_scatter_iiiR, 
         panel_height = PANEL_H * 0.9, panel_width = PANEL_W * 0.9 )
```

## Fig 1e
```{r}
df <- df %>%
  mutate(S5 = case_when(
    insert_5 == "eNMU_region_iii_forward"            ~ "e-NMU(iii)F",
    insert_5 == "eNMU_region_iii_reverse"             ~ "e-NMU(iii)R",
    insert_5 == "eNMU_region_iv_reverse"              ~ "e-NMU(iv)R",
    str_detect(insert_5, "^synthetic_insulator")      ~ "Insulator",
    str_detect(insert_5, "^dinucleotide_shuffled")    ~ "Controls",
    str_detect(insert_5, "^eNMU_region")              ~ "Other e-NMU",
    TRUE                                               ~ "Other"
  ),
  S5 = factor(S5, levels = c("e-NMU(iii)F", "e-NMU(iii)R", "e-NMU(iv)R",
                              "Other e-NMU", "Controls", "Insulator", "Other")))


# Compute n per S5 group, build two-line labels: "S5 name\nn = count"
n_labels <- df %>%
  count(S5) %>%
  mutate(label = paste0(S5, "\nn = ", scales::comma(n)))

# Named vector for scale_x_discrete(labels = ...), keyed by factor level
label_map <- setNames(n_labels$label, n_labels$S5)

p <- ggplot(df, aes(x = S5, y = avg_Rep + 0.05, fill = S5)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, aes(fill = S5), outlier.shape = NA) +
  scale_fill_manual(values = c(
    "e-NMU(iii)F" = unname(COLOR_IIIF),
    "e-NMU(iii)R" = unname(COLOR_IIIR),
    "e-NMU(iv)R"  = unname(COLOR_IVRF),
    "Other e-NMU" = COLOR_ENMU_OTHER,
    "Controls"    = COLOR_CTRL_ALL,
    "Insulator"   = COLOR_INSU_ALL,
    "Other"        = unname(COLOR_OTHER)
  )) +
  scale_x_discrete(labels = label_map) +
  log_scale("y", expand = expansion(mult = c(0.05, 0.25)), breaks = 10^(-2:4)) +
  annotation_logticks(sides = "l", size = 0.3) +
  labs(
    x = "S5 identity",
    y = "Activity score"
  ) +
  theme(
    legend.position  = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, vjust = 1, margin = margin(t = 2))
  )

save_fig("fig1e_MPRA_score_violin", p, panel_height = PANEL_H * 0.9, panel_width = PANEL_W * 1.3)
print(p)
```


# Figure S1
## Fig S1b
```{r figS1b-insert-counts-bc-level}
# Prepare plot data from table2_bc_counts (computed in pacbio-association-prep)
plot_data_bc <- table2_bc_counts %>%
  group_by(base_name, orientation) %>%
  summarise(total_counts = sum(total_counts), .groups = "drop") %>%
  pivot_wider(names_from = orientation, values_from = total_counts, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total)) %>%
  dplyr::select(-Total)

# Convert to long format
plot_long_bc <- plot_data_bc %>%
  pivot_longer(cols = -base_name, names_to = "Orientation", values_to = "Count")

# Plot MODIFIED THIS TO ADJUST THE NAMES USING SHORT_LABEL()
insert_order_labels <- c("Insu(i)",
  "e-NMU(i)", "e-NMU(ii)", "e-NMU(iii)", "e-NMU(iv)", "e-NMU(v)",
  "Ctrl(i)",  "Ctrl(ii)",  "Ctrl(iii)",  "Ctrl(iv)",  "Ctrl(v)")

plt_insert_counts_bc <- plot_long_bc %>%
  mutate(insert_label = short_label(paste0("insert_1_id_", base_name)) %>%
           str_remove("-S\\d+$"),
         insert_label = factor(insert_label, levels = insert_order_labels)) %>%
  ggplot(aes(x = insert_label, y = Count, fill = Orientation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +
  scale_y_continuous(
    limits = c(0, 5e6),
    breaks = c(0, 1e6, 2e6, 3e6, 4e6, 5e6),
    labels = c("0", expression(1%*%10^6), expression(2%*%10^6),
               expression(3%*%10^6), expression(4%*%10^6), expression(5%*%10^6))
  ) +
  labs(x = "Element", y = "# of reads", fill = "Orientation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
                                   margin = margin(t = 2)),
                legend.text  = element_text(size = 9), 
                 legend.title = element_text(size = 9))  

print(plt_insert_counts_bc)

save_fig("figS1b_insert_counts_bc_level", plt_insert_counts_bc,
  panel_height = PANEL_H * 0.9, panel_width = PANEL_W * 1.1)
```

## Fig S1c
```{r figS1c-stacked-detail}

SLOTS <- 1:5
association_df_filter <- association_df %>% dplyr::select(!insert_combo)

# ── Build detail data ─────────────────────────────────────────────────────────
slot_element_detail <- map_dfr(SLOTS, function(s) {
  col <- paste0("insert_", s)
  if (!col %in% names(association_df_filter)) return(NULL)
  tibble(slot = s,
         element_value = association_df_filter[[col]],
         count = association_df_filter$count) %>%
    mutate(insert_type = str_remove(element_value, "_(forward|reverse)$"))
}) %>%
  group_by(slot, insert_type) %>%
  summarise(n_present = sum(count), .groups = "drop") %>%
  group_by(slot) %>%
  mutate(
    pct         = n_present / sum(n_present) * 100,
    fill_color  = coalesce(color_map_detail[insert_type], "#AAAAAA"),
    insert_label = factor(
      type_order_labels[match(insert_type, type_order_types)],
      levels = type_order_labels
    )
  ) %>%
  filter(!is.na(insert_label))   # drop any unmapped types

# Legend built from type_order directly so all 11 types always appear,
# even if absent from this particular dataset
legend_df <- tibble(
  insert_type  = type_order_types,
  insert_label = factor(type_order_labels, levels = type_order_labels),
  fill_color   = color_map_detail[type_order_types]
)

# ── Plot ──────────────────────────────────────────────────────────────────────
p_stacked_detail <- ggplot(slot_element_detail,
                           aes(x = factor(slot), y = pct,
                               fill = fill_color,
                               group = insert_label)) +
  geom_col(width = 0.6, color = "white", linewidth = 0.3) +
  scale_x_discrete(labels = c("1" = "S1", "2" = "S2", "3" = "S3", "4" = "S4", "5" = "S5")) +
  scale_fill_identity(
    guide  = "legend", name = NULL,
    breaks = legend_df$fill_color,
    labels = as.character(legend_df$insert_label)
  ) +
  labs(x = "Slot", y = "% of reads") +
  theme(
    legend.position  = "right",
    legend.text      = element_text(size = 9),
    axis.title.x     = element_text(margin = margin(t = 7))  # adjust t as needed
  )

p_stacked_detail
save_fig("figS1c_stacked_detail", p_stacked_detail, panel_height = PANEL_H * 0.9, panel_width = PANEL_W * 1.1)
```

## Fig S1d
```{r}
# ── All-22-insert violin, sorted by mean activity (descending), no n= labels ──
df <- df %>%
  mutate(
    insert_type_S5 = str_remove(insert_5, "_(forward|reverse)$"),
    orientation_S5 = case_when(
      str_detect(insert_5, "_forward$") ~ "F",
      str_detect(insert_5, "_reverse$") ~ "R",
      TRUE ~ NA_character_
    ),
    base_label_S5  = type_order_labels[match(insert_type_S5, type_order_types)],
    S5_all         = paste0(base_label_S5, orientation_S5)
  ) %>%
  filter(!is.na(S5_all))

# Order levels left -> right by descending mean log-activity
S5_order <- df %>%
  group_by(S5_all) %>%
  summarise(mean_act = mean(avg_Rep_log05, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_act)) %>%
  pull(S5_all)

df <- df %>%
  mutate(S5_all = factor(S5_all, levels = S5_order))

# One color per base type (i-v / Ctrl i-v / Insu), shared across F and R
fill_map_base <- setNames(color_map_detail[type_order_types], type_order_labels)
fill_vals_all <- setNames(
  fill_map_base[sub("[FR]$", "", S5_order)],
  S5_order
)

p_all22 <- ggplot(df, aes(x = S5_all, y = avg_Rep + 0.05, fill = S5_all)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, aes(fill = S5_all), outlier.shape = NA) +
  scale_fill_manual(values = fill_vals_all) +
  log_scale("y", expand = expansion(mult = c(0.05, 0.25)), breaks = 10^(-2:4)) +
  annotation_logticks(sides = "l", size = 0.3) +
  labs(
    x = "S5 identity",
    y = "Activity score"
  ) +
  theme(
    legend.position  = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1, vjust = 1, margin = margin(t = 2))
  )

print(p_all22)
save_fig("figS1d_all22_insert_violin", p_all22, panel_height = PANEL_H * 0.9, panel_width = PANEL_W * 3.4)
```
